### 2022-02-02 paper/manuscript.tex
Initial commit

%%
%% This is file `sample-lualatex.tex', % generated with the docstrip utility.
%%
%% The original source files were:
%%
%% samples.dtx  (with options: `sigconf')
%% 
%% IMPORTANT NOTICE:
%% 
%% For the copyright see the source file.
%% 
%% Any modified versions of this file must be renamed % with new filenames
%distinct from sample-lualatex.tex.
%% 
%% For distribution of the original source see the terms % for copying and
%modification in the file samples.dtx.
%% 
%% This generated file may be distributed as long as the % original source
%files, as listed above, are part of the % same distribution. (The sources need
%not necessarily be % in the same archive or directory.)
%%
%%
%% Commands for TeXCount TC:macro \cite [option:text,text] TC:macro \citep
%[option:text,text] TC:macro \citet [option:text,text] TC:envir table 0 1
%TC:envir table* 0 1 TC:envir tabular [ignore] word TC:envir displaymath 0 word
%TC:envir math 0 word TC:envir comment 0 0
%%
%%
%% The first command in your LaTeX source must be the \documentclass command.
\documentclass[sigconf, anonymous, review]{acmart}
\usepackage{probs}
\usepackage[linesnumbered,ruled]{algorithm2e}
\mathchardef\mhyphen="2D
%%
%% \BibTeX command to typeset BibTeX logo in the docs
\AtBeginDocument{%
  \providecommand\BibTeX{{%
    \normalfont B\kern-0.5em{\scshape i\kern-0.25em b}\kern-0.8em\TeX}}}

%% Rights management information.  This information is sent to you % when you
%complete the rights form.  These commands have SAMPLE % values in them; it is
%your responsibility as an author to replace % the commands and values with
%those provided to you when you % complete the rights form.
\setcopyright{acmcopyright}
\copyrightyear{2022}
\acmYear{2022}
%\acmDOI{10.1145/1122445.1122456}
\acmConference[KDD'22]{the 28th ACM SIGKDD Conference on Knowledge Discovery and
Data Mining}{August 2022}{Washington, DC, USA}

%% These commands are for a PROCEEDINGS abstract or paper.
%\acmConference[Woodstock '18]{Woodstock '18: ACM Symposium on Neural Gaze
%Detection}{June 03--05, 2018}{Woodstock, NY} \acmBooktitle{Woodstock '18: ACM
%Symposium on Neural Gaze Detection, June 03--05, 2018, Woodstock, NY}
%\acmPrice{15.00} \acmISBN{978-1-4503-XXXX-X/18/06}

%%
%% Submission ID. % Use this when submitting an article to a sponsored event.
%You'll % receive a unique submission ID from the organizers % of the event, and
%this ID should be used as the parameter to this command.
%%\acmSubmissionID{123-A56-BU3}

%%
%% The majority of ACM publications use numbered citations and % references.
%The command \citestyle{authoryear} switches to the % "author year" style.
%%
%% If you are preparing content for an event % sponsored by ACM SIGGRAPH, you
%must use the "author year" style of % citations and references. % Uncommenting
%% the next command will enable that style. %\citestyle{acmauthoryear}

%%
%% end of the preamble, start of the body of the document source.
\begin{document}

%%
%% The "title" command has an optional parameter, % allowing the author to
%define a "short title" to be used in page headers.
\title{Predictive State Geometry via Cantor Embeddings and Wasserstein Distance}
%%
%% The "author" command and its associated commands are used to define % the
%authors and their affiliations. % Of note is the shared affiliation of the
%first two authors, and the % "authornote" and "authornotemark" commands % used
%to denote shared contribution to the research.
\author{Samuel P. Loomis}
\email{sloomis@ucdavis.edu}
\orcid{0000-0003-1587-9874}

\author{James P. Crutchfield}
\email{chaos@ucdavis.edu}
\orcid{0000-0003-4466-5410}
\affiliation{%
  \institution{Complexity Sciences Center and Physics Department, \\University of California at Davis}
  \streetaddress{One Shields Ave.}
  \city{Davis}
  \state{CA}
  \country{USA}
  \postcode{95616}
}

%%
%% By default, the full list of authors will be used in the page % headers.
%Often, this list is too long, and will overlap % other information printed in
%the page headers. This command allows % the author to define a more concise
%list % of authors' names for this purpose.
\renewcommand{\shortauthors}{Loomis and Crutchfield}

%%
%% The abstract is a short summary of the work to be presented in the % article.
\begin{abstract}
  Predictive states for stochastic processes are a non-parametric and
  interpretable construct with relevance across a multitude of modeling
  paradigms. Recent progress on the self-supervised reconstruction of predictive
  states from time-series data has focused on the use of reproducing kernel
  Hilbert spaces. Here, we examine how Wasserstein distances may be used to
  detect predictive equivalences in symbolic data. We construct a
  finite-dimensional embedding of sequences based on the Cantor set, and use
  distances in this embedding to compute Wasserstein distances between
  distributions over sequences (``predictions''). We show that analyzing the
  resulting geometry (via hierarchical clustering and dimension reduction)
  provides insight into the temporal structure of processes ranging from the
  relatively simple (\emph{e.g.}\ hidden Markov models) to the very complex
  (\emph{e.g.}\ indexed grammars).
\end{abstract}

%%
%% The code below is generated by the tool at http://dl.acm.org/ccs.cfm. %
%Please copy and paste the code instead of the example below.
%%
\begin{CCSXML}
  <ccs2012>
  <concept>
  <concept_id>10002950.10003648.10003649</concept_id>
  <concept_desc>Mathematics of computing~Probabilistic representations</concept_desc>
  <concept_significance>500</concept_significance>
  </concept>
  <concept>
  <concept_id>10002950.10003648.10003649.10003657</concept_id>
  <concept_desc>Mathematics of computing~Nonparametric representations</concept_desc>
  <concept_significance>500</concept_significance>
  </concept>
  <concept>
  <concept_id>10002950.10003648.10003700.10003701</concept_id>
  <concept_desc>Mathematics of computing~Markov processes</concept_desc>
  <concept_significance>500</concept_significance>
  </concept>
  <concept>
  <concept_id>10002950.10003648.10003700</concept_id>
  <concept_desc>Mathematics of computing~Stochastic processes</concept_desc>
  <concept_significance>500</concept_significance>
  </concept>
  <concept>
  <concept_id>10002950.10003648.10003688.10003693</concept_id>
  <concept_desc>Mathematics of computing~Time series analysis</concept_desc>
  <concept_significance>500</concept_significance>
  </concept>
  <concept>
  <concept_id>10002950.10003741.10003742</concept_id>
  <concept_desc>Mathematics of computing~Topology</concept_desc>
  <concept_significance>300</concept_significance>
  </concept>
  <concept>
  <concept_id>10002950.10003712</concept_id>
  <concept_desc>Mathematics of computing~Information theory</concept_desc>
  <concept_significance>300</concept_significance>
  </concept>
  <concept>
  <concept_id>10003752.10003766</concept_id>
  <concept_desc>Theory of computation~Formal languages and automata theory</concept_desc>
  <concept_significance>300</concept_significance>
  </concept>
  <concept>
  <concept_id>10003752.10003766.10003771</concept_id>
  <concept_desc>Theory of computation~Grammars and context-free languages</concept_desc>
  <concept_significance>300</concept_significance>
  </concept>
  </ccs2012>
\end{CCSXML}

\ccsdesc[500]{Mathematics of computing~Probabilistic representations}
\ccsdesc[500]{Mathematics of computing~Nonparametric representations}
\ccsdesc[500]{Mathematics of computing~Stochastic processes}
\ccsdesc[500]{Mathematics of computing~Time series analysis}
\ccsdesc[300]{Mathematics of computing~Markov processes}
\ccsdesc[300]{Mathematics of computing~Topology}
\ccsdesc[300]{Mathematics of computing~Information theory}
\ccsdesc[300]{Theory of computation~Formal languages and automata theory}
\ccsdesc[300]{Theory of computation~Grammars and context-free languages}

%%
%% Keywords. The author(s) should pick words that accurately describe % the work
%being presented. Separate the keywords with commas.
\keywords{time series, predictive states, wasserstein distance, hierarchical clustering}

%% A "teaser" image appears between the author and affiliation % information and
%the body of the document, and typically spans the % page.

%%
%% This command processes the author and affiliation and title % information and
%builds the first part of the formatted document.
\maketitle

\section{Introduction}
Suppose that we have a finite sequence $x_1 \dots x_L$ of categorical
observations drawn from a temporal process. We may suppose that the process is
stationary (time-translation-invariant) and ergodic (has a tendency to explore
all possible behaviors) \cite{Kall01a,Shal07a}. We may wish to forecast from the
observed information the behavior of the next $n$ observations. If we suspect
that temporal correlations of the process do not matter much beyond $k$ symbols,
then we can use the data we have to take all subsequences of length $n+k$,
splitting each subsequence into words of length $k$ and $n$ respectively. From
these ``past/future'' pairs we could construct an empirical conditional
distribution
\begin{equation}\label{eq:conditional}
  \hat{P}_{x_{1}\dots x_{n}|x_{-k+1}\dots x_0} = 
  \frac{C_{x_{-k+1}\dots x_{n}}}{C_{x_{-k+1}\dots x_0}}
\end{equation}
where $C_{w}$ is the number of times the word $w$ appears in our sequence $x_1
\dots x_L$.

But suppose now that we do not know how long-range the temporal dependencies in
our process can stretch. Even very simple stochastic processes can have infinite
Markov order, indicating potential long-term dependence of future observations
on the past \cite{Uppe97a}. Given sufficient data, it would be desirable to take
pasts of arbitrary length, and converge towards some prediction conditioned on
the \emph{infinite} past:
\begin{equation}\label{eq:convergence}
  {P}_{x_{1}\dots x_{n}|\overleftarrow{x}}
  = \lim_{k\rightarrow\infty}
  \hat{P}_{x_{1}\dots x_{n}|x_{-k}\dots x_0}
\end{equation}
where $\overleftarrow{x}$ denotes an infinite sequence $\overleftarrow{x} =
(\dots,x_{-1},x_0)$ of observations stretching into the past. This mathematical
ideal is known to previous literature as the \emph{causal} or \emph{predictive
state} \cite{Crut88a,Jaeg00a}. Formally, the conditional predictions
${P}_{x_{1}\dots x_{n}|\overleftarrow{x}}$ for all forecast lengths $n$ together
describe a \emph{probability measure} over future sequences $\overrightarrow{x}
= (x_1,x_2,\dots)$; the predictive state is this measure. We will denote it
simply by $P_{\overleftarrow{x}}$.

Predictive states have been utilized for inference and modeling in dynamical
systems \cite{Crut92c,Emen16a,Salo19a}, renewal processes and spike-trains
\cite{Marz14e,Marz15a}, condensed matter physics \cite{Varn12a,Varn14a} and
spatiotemporal systems \cite{Rupe19a,Rupe19b,Rupe20a}, and a deep mathematical
theory of predictive state inference has been correspondingly developed
\cite{Jaeg00a,Shal02a,Jame04a,Stil07b,Stre13a,Thon15a,Marz19a,Brod20a}. Despite
this, universal conditions for predictive state convergence were not given until
recently. It is now known that if the dataset $x_1 \dots x_L$ is drawn from any
stationary process, and if $L$ is sufficiently large, then Eq.
\eqref{eq:convergence} will converge for any $n$ and a probability-1 subset of
pasts $\overleftarrow{x}$ \cite{Loom21a}. In the language of measures,
$P_{\overleftarrow{x}}$ converges \emph{in distribution}, with respect to the
\emph{product topology} of the space of sequences $\mathcal{X}^\mathbb{N}$.

The general goals of predictive state analysis are typically threefold
\cite{Shal01a}. The first is to understand the overall structure of how the
predictive states relate to one another geometrically, and possibly use this
geometry to classify pasts based on equivalence of their predictive states. The
second is to actually reproduce the prediction to a specified accuracy. The
third is to understand the dynamics of how predictions evolve under a stream of
new observations. We will focus in this paper on the first, as it is a crucial
building block to achieving the other two.

Recent attempts at reconstructing the geometry of predictive states have focused
on embedding them in reproducing kernel Hilbert spaces
\cite{Song09a,Song10a,Boot13a,Brod20a,Loom21a}, to great effect largely because
convergence in Hilbert spaces generated by universal kernels is equivalent to
convergence in distribution \cite{Srip10a}. That is, these embeddings allow
accurate representations of predictive states because they respect the product
topology of sequences in the same way that predictive states themselves do.

Understanding the source of the power of RKHS methods in predictive state
analysis frees us to consider other options. In this paper, we will embed
symbolic sequences in a one-dimensional space that has the same topology as the
product space. This embedding is inspired by the fractal Cantor set
\cite{Kurk03a}. Predictive states can then be thought of as distributions in
this one-dimensional space. We then use the Wasserstein distance to compute the
geometry between predictive states, which can be computed with a closed-form
integral for one-dimensional distributions. This works as an alternative to
RKHS-based distances because the Wasserstein distance also reproduces the
topology of convergence in distribution \cite{Pane19a}. The resulting distance
matrix can be used to find low-dimensional embeddings \cite{Borg05a} of the
geometry or hierarchical clusterings \cite{Mull11a} of the predictive states.
The latter, in particular, when combined with our fractal embedding, provides a
highly interpretable visualization of the predictive state space.

\section{Example processes}
The methods in this paper are intended to be applied toward stationary and
ergodic stochastic processes which produce categorical time-series data. For our
intents we can think of a stochastic process as a collection of probability
distributions $\Prob{\mu}{x_1\dots x_L}$ over any finite, contiguous sequence,
taking values in a finite set $\mathcal{X}$. Formally this describes a measure
$\mu$ over the set of all bi-infinite sequences $(\dots,x_{-1},x_0,x_1,\dots)\in
\mathcal{X}^{\mathbb{Z}}$. These sorts of processes can be generated by a number
of systems with widely varying complexity. Most popularly studied are those
methods often characterized as having some degree of ``finite memory'': Markov
models, hidden Markov models, and observable operator models (also termed
generalized hidden Markov models) \cite{Uppe97a,Jaeg00a}. Beyond these one can
also generate processes using probabilistic grammars, such as probabilistic
context-free and indexed grammars \cite{Gema00a}. Additionally, coarse-grained
data from dynamical systems (such as the logistic map) can display behavior
varying widely in complexity \cite{Crut92c}. In this paper we will refer back
frequently to the following example processes:
\begin{enumerate}
  \item The \emph{even process} can be generated by repeatedly tossing a coin
  and writing down a $0$ for every tail and $11$ for every head. The process is
  essentially random except for the constraint that $1$'s can only appear in
  contiguous blocks of even size. The even process has infinite Markov order but
  can be generated by a two-state hidden Markov process \cite{Crut01a}.
  \item The $a^n b^n$ \emph{process} can be generated by choosing a random
  integer $n\geq 1$ (we will suppose via a Poisson process) and writing $n$
  $\mathtt{a}$'s followed by an equal number of $\mathtt{b}$'s, and then
  repeating this process indefinitely. The result will be sequences where any
  contiguous block of $\mathtt{a}$'s is followed by a block of $\mathtt{b}$'s of
  equal size. The $\mathtt{a}^n \mathtt{b}^n$ process cannot be modeled by any
  hidden Markov model, though it is a simple example of a probabilistic
  context-free grammar \cite{Hopc06a}.
  \item The $x+f(x)$ \emph{process} is a probabilistic context-free grammar
  modeling simple mathematical expressions. It has terminal symbols
  $\{\mathtt{(}\ ,\ \mathtt{)}\ ,\ \mathtt{;}\ ,\ \mathtt{+}\ ,\ \mathtt{f}\ ,\
  \mathtt{x}\}$ and non-terminals $\{A,B,C\}$, and starts with a sequence osf
  $A$'s with the production rules:
  \begin{align*}
    A &\mapsto B\ \mathtt{+}\ C\ \mathtt{;}\ |\ C\ \mathtt{;}\\
    B &\mapsto B\ \mathtt{+}\ C\ |\ C\\
    C &\mapsto \mathtt{f(}B\mathtt{)}\ |\ \mathtt{x}
  \end{align*}
  \item The $a^n b^n c^n$ \emph{process} is a probabilistic indexed grammar
  \cite{Hopc06a} which is analagous to $\mathtt{a}^n \mathtt{b}^n$ except after
  writing the blocks of $\mathtt{a}$'s and $\mathtt{b}$'s, we also write a block
  of $\mathtt{c}$'s of length $n$.
  \item The \emph{Feigenbaum process} is generated by sampling from the time
  series of the logistic map at critical parameter $r\approx 3.56995$:
  \begin{equation*}
    y_{t+1} = r y_t(1-y_t)
  \end{equation*}
  and then coarse-graining the data by taking $x_t = 0$ if $0< y_t \leq
  \frac{1}{2}$ and $x_t = 1$ if $\frac{1}{2} < y_t <1$ \cite{Kurk03a}.
  Alternatively, we can generate this process by starting with a single $0$ and
  executing the replacements $0\mapsto 11$ and $1 \mapsto 01$ consecutively. The
  result of the Feigenbaum process is indexed-context free \cite{Crut92c}.
\end{enumerate}

\section{Cantor-embedding sequences}
\begin{figure*}[h]
  \centering
  \includegraphics[width=0.4\linewidth]{../plots/even_CANTOR.png}
  \includegraphics[width=0.4\linewidth]{../plots/anbn_CANTOR.png}
  \includegraphics[width=0.4\linewidth]{../plots/anbncn_CANTOR.png}
  \includegraphics[width=0.4\linewidth]{../plots/x+f(x)_CANTOR.png}
  \caption{ Cantor plots for the even, $\mathtt{a}^n\mathtt{b}^n$,
    $\mathtt{a}^n\mathtt{b}^n\mathtt{c}^n$, and $\mathtt{x+f(x)}$ processes.
    Each point $(x,y)$ corresponds to a pair of sequences corresponding to the
    past and future respectively. The symbol on the $x$ ($y$) axis indicate that
    all points above (to the right of) that symbol have a past (future) whose
    most recent observation is that symbol. Though not marked, further
    proportional subdivisions of each segment of the axes would indicate the
    value of the second, third, \emph{etc.} symbols. For instance, one can read
    from the $\mathtt{x+f(x)}$ fractal that any past which ends in $\mathtt{f}$
    must be paired with a future beginning in $\mathtt{(f}$ or $\mathtt{(x}$. }
    \label{fig:cantor}
  \Description{Four fractals are displayed, respectively labeled as the even,
  $\mathtt{a}^n\mathtt{b}^n$, $\mathtt{a}^n\mathtt{b}^n\mathtt{c}^n$, and
  $\mathtt{x+f(x)}$ processes.}
\end{figure*}

The geometry of sequences is inherently self-similar. Given an infinite sequence
$\overrightarrow{x} = (x_1,x_2,\dots)$, we can split it into its leading word
$x_1 x_2 \dots x_L$ and a following sequence $\overrightarrow{x}_L =
(x_L,x_{L+1},\dots)$. That is, the space of sequences $\mathcal{X}^\mathbb{N}$
can be factored into $\mathcal{X}^L\times \mathcal{X}^\mathbb{N}$ for any $L$.
The fractal nature of sequence-space is encoded in the structure of its
\emph{product topology}.

We can exploit this self-similarity in an interesting way by constructing a
mapping between sequence space and the celebrated Cantor set (or one of its
generalizations). Suppose a symbolic sequence $(x_1,x_2,\dots)$ takes values in
an alphabet $\mathcal{X}$ of size $|\mathcal{X}|$. To each $x\in\mathcal{X}$ we
can associate some unique integer between $0$ and $|\mathcal{X}|-1$ inclusive;
call this $J(x)$. Then there is a function $C:\mathcal{X}^\mathbb{N}\rightarrow
[0,1]$ which maps every sequence to a positive real number:
\begin{equation}
  C(x_1,x_2,\dots) = \sum_{k=1}^\infty \frac{2J(x_k)}{(2|\mathcal{X}|-1)^k}
\end{equation}
For instance, suppose that $|\mathcal{X}|=2$ has two elements; then $C$ maps the
sequence to a point the traditional Cantor set fractal. For a finite sequence of
length $L$, truncate the sum at $k=L$.

Remarkably, the embedding $C$ has the property that for any continuous function
$f$ on $[0,1]$, the function $F(\overrightarrow{x}) = f(C(\overrightarrow{x}))$
is continuous on $\mathcal{X}^\mathbb{N}$; further, if $F$ is continuous on
$\mathcal{X}^\mathbb{N}$, then $f(y) = F(C^{-1}(y))$ is continuous on the image.
Thus the embedding $C$ respects the basic structure of the product topology
\cite{Kurk03a}.

Stationary processes, due to their time-translation invariance, inherit the
fractal temporality of sequence space. This can be neatly visualized: given a
length-$L$ sample $x_1\dots x_L$, and some $n,k>0$, take a sliding window of
pasts and futures, $(x_{t-n+1}\dots x_t, x_{t+1}\dots x_k)$ for $t=n,\dots,L-k$.
For each past-future pair, compute the truncated Cantor embeddings on the
\emph{reversed} past and (unreversed) future: $\left(C(x_t \dots
x_{t-n+1}),C(x_{t+1}\dots x_k)\right)$. The resulting pairs of real numbers can
be plotted as $(x,y)$-values on a scatter plot. The fractal which emerges
contains, in essence, all information necessary to understand the temporal
structures of the process. See Fig. \ref{fig:cantor} for examples and guidance
on how to interpret the visualization.

It is worth noting that for $|\mathcal{X}|>2$, the embedding $C$ also introduces
additional structure which may or may not be desirable. The association of each
symbol $x$ with an integer $j_x$ endows an ordinal structure on the set
$\mathcal{X}$; this ordinality is present in the macroscopic geometry of
$C\left(\mathcal{X}^{\mathbb{N}}\right)$. Later we will examine higher-dimension
embeddings which do not have this ordinal aspect; however, they will come at the
cost of increased computational complexity for the Wasserstein distance
computation, so for now we will assume that ordinal artifacts are either desired
or sufficiently tolerable as to not outweight the computational benefits of
working in one dimension.

\section{Wasserstein distance on predictive states}

\begin{algorithm}[t]
  \SetKwData{UnqPasts}{UnqPasts}\SetKwData{Cantors}{Cantors}\SetKwData{Wass}{Wass}
  \SetKwFunction{Wasserstein}{Wasserstein}\SetKwFunction{Append}{append}
  \SetKwFunction{Index}{index} \SetKwFunction{Length}{length}
  \SetKwFunction{Matrix}{Matrix}
  \SetKwFunction{CantorWasserstein}{CantorWasserstein}
  \SetKwInOut{Input}{input}\SetKwInOut{Output}{output}

  \underline{function \CantorWasserstein} ($n,k,x_1\dots x_L$)\;
  \Input{Integers $n,k$ of past and future lengths} 
  \Input{Length-$L$ sequence $x_1\dots x_L$ of observations} 
  \Output{List $\UnqPasts$ of unique pasts}
  \Output{List of lists $\Cantors$ of Cantor-embedded futures} 
  \Output{Matrix $\Wass$ of Wasserstein distances} 
  $\UnqPasts \leftarrow \left[\right]$\; 
  $\Cantors \leftarrow \left[\right]$\; 
  \For{$t\leftarrow n$ \KwTo $L-k$} { 
    $\overleftarrow{\mathbf{x}} \leftarrow \left[x_{t-n+1},\dots, x_{t}\right]$\;
    $\overrightarrow{\mathbf{x}} \leftarrow \left[x_{t+1},\dots, x_{t+k}\right]$\;
      \vspace{-1.5em}
      \begin{flalign*}
        p &\leftarrow \sum_{\ell=1}^L
        \frac{2J(\overrightarrow{x}_\ell)}{(2|\mathcal{X}|-1)^\ell}\ ; &&
      \end{flalign*}

      \eIf{$\overleftarrow{\mathbf{x}} \in \UnqPasts$} { 
        $\Append\ \overleftarrow{\mathbf{x}}$ \KwTo $\UnqPasts$\; 
        $\Append\ [p]$ \KwTo $\Cantors$\; 
        } { 
        $j \leftarrow \Index\left(\overleftarrow{\mathbf{x}},\UnqPasts\right)$\;
        $\Append\ p$ \KwTo $\Cantors_j$\; 
        } 
  }
  $K \leftarrow \Length(\UnqPasts)$\; 
  $\Wass \leftarrow \Matrix(K,K)$\;
  \For{$k\leftarrow 1$ \KwTo $K$} { 
    \For{$\ell\leftarrow 1$ \KwTo $K$} {  
      $\Wass_{k \ell}
        \leftarrow
        \Wasserstein(\Cantors_k,\Cantors_\ell)$\;
        $\Wass_{\ell k} \leftarrow \Wass_{k \ell}$\; } } 
  \KwResult{$\UnqPasts$,$\Cantors$,$\Wass$}
  \caption{Steps for converting a sequence of categorical time-series data
        into a labeled collection of empirical distributions of Cantor-embedded futures,
        and a matrix of Wasserstein distances from said distributions.}
  \label{alg:cantorwass}
\end{algorithm}

The Cantor fractals in Fig. \ref{fig:cantor} can be thought of as probability
distributions. We can consequently interpret a vertical slices of the fractal,
located at horizontal position $C(\overleftarrow{x})$, as visualizing the
predictive state $P_{\overleftarrow{x}}$ as a distribution over Cantor-embedded
futures, $C(\overrightarrow{x})$.

For example, by analyzing the Cantor fractal of the even process, one may notice
that there are effectively only 2 distinct predictive states---every vertical
column is just one of two types. This corresponds with the 2 states of the
hidden Markov model which generates the even process.

This visualization allows us to see how predictive states distribute their
probability over the intrinsic geometry of potential futures. We can compare
predictive states not only on how much their supports overlap, but on how
geometrically close their supports are to one another. For the $\mathtt{a}^n
\mathtt{b}^n$ process, for example, we can see that the first few columns
(corresponding to pasts of the form $\dots \mathtt{b a}^n$ for some $n$) are
inherently similar to one another, though they are shifted upwards the closer to
the axis they are (which corresponds to the increasing number of $\mathtt{b}$'s
in the predicted future as $n$ increases).

The intuitive distance metric between probability measures for capturing this
underlying geometry is the Wasserstein metric \cite{Pane19a}. Given two measures
$\mu$, $\nu$ defined on a metric space $\mathcal{M}$ with metric $d$, the
Wasserstein distance between $\mu$ and $\nu$ is given by
\begin{equation*}
  W(\mu,\nu) = \min_{\pi\in \Gamma(\mu,\nu)}
   \int_{\mathcal{M}\times\mathcal{M}} 
   d(x,y) d\pi(x,y)
\end{equation*}
where $\Gamma(\mu,\nu)$ is the set of all measures on
$\mathcal{M}\times\mathcal{M}$ whose left and right marginals are $\mu$ and
$\nu$ respectively. This can be interpreted as the minimal cost for ``shifting''
the probability mass from one distribution to match shape of the other.

$W(\mu,\nu)$ the solution to a constrained linear optimization. As a function of
distributions, $W(\mu,\nu)$ is continuous with respect to convergence in
distribution; in fact, convergence under the Wasserstein distance is equivalent
to convergence in distribution on compact spaces. This makes $W(\mu,\nu)$ ideal
for measuring geometry between predictive states, since empirical estimates of
these are known to converge in distribution.

When $\mathcal{M} \subseteq \mathbb{R}$, there is in fact a closed-form solution
to the Wasserstein optimization problem \cite{Thas10a}. Let $F$ and $G$
respectively be the cumulative distributions functions of $\mu$ and $\nu$. Then
\begin{equation}
  W(\mu,\nu) = \int_{-\infty}^\infty \left|F(t)-G(t)\right| dt
\end{equation}
This closed-form solution is considerably faster to compute than the linear
optimization required for arbitrary metric spaces. Because the Cantor embedding
embeds the space of sequences directly into $[0,1]$, we can utilize this
formula.

The combination of the Cantor embedding and the Wasserstein distance lays out a
fairly straightforward programme for analyizing a categorical time series:
\begin{enumerate}
  \item Apply Algorithm \ref{alg:cantorwass} to the data stream $x_1\dots x_L$
  for a specified past length $k$ and future length $n$, retrieving the set of
  unique observed pasts, the empirical Cantor distributions corresponding to
  each past, and the matrix of Wasserstein distances computed from these
  distributions.
  \item Use the Wasserstein distance matrix to perform additional methods of
  geometric data analysis, such as hierarchical clustering \cite{Mull11a} and
  multidimensional scaling \cite{Borg05a}, in order to elucidate the relative
  geometry of the predictive states.
\end{enumerate}
In the next two sections we examine the results of this approach.

\section{Interpretable predictive states with hierarchical clustering}
\begin{figure*}[ht]
  \centering
  \includegraphics[width=0.45\linewidth]{../plots/even_HCLUST.png}
  \includegraphics[width=0.45\linewidth]{../plots/anbn_HCLUST.png}
  \includegraphics[width=0.45\linewidth]{../plots/anbncn_HCLUST.png}
  \includegraphics[width=0.45\linewidth]{../plots/x+f(x)_HCLUST.png}
  \caption{From upper left to lower right, the clustered Cantor diagrams of the even, $\mathtt{a}^n\mathtt{b}^n$,
  $\mathtt{a}^n\mathtt{b}^n\mathtt{c}^n$, and $\mathtt{x+f(x)}$ processes (zoom
  for detail). On
  each diagram, the vertical axis shows all pasts of a given length (length
  $k=8$ for the even, $\mathtt{a}^n\mathtt{b}^n$ and
  $\mathtt{a}^n\mathtt{b}^n\mathtt{c}^n$ processes and $k=4$ for the
  $\mathtt{x+f(x)}$ process), along with their hierarchically clustered
  dendrogram. For the purposes of this graphic the coloring threshold was
  chosen to aid in visual interpretation. The lines in each row show the
  empirical distribution of Cantor-embedded futures observed following each
  past. As such the horizontal axis corresponds exactly to the vertical axis of
  Fig. \ref{fig:cantor}.}
    \label{fig:hclust}
  \Description{Four graphs are displayed. Each vertical axis has a list of symbolic sequences
  organized by a hierarchical clustering tree. The horizontal axis is a list of
  individual symbols. The content of each graph is rows of fractally-distributed
  lines.}
\end{figure*}
In Fig. \ref{fig:hclust} we display the result of collecting the Cantor-embedded
empirical predictions for all pasts of a given length, for four processes (even,
$\mathtt{a}^n \mathtt{b}^n$, $\mathtt{a}^n \mathtt{b}^n \mathtt{c}^n$ and
$\mathtt{x+f(x)}$). In each case the Wasserstein distance between every pair of
predictions was computed and used to hierarchically cluster the pasts (using the
Ward method \cite{Mull11a}) with
others that produced similar predictions.

The resulting clustered Cantor plots offer a highly interpretable visualization
of the relationship between pasts and futures, and of the geometry of the
predictive states. They are, in a certain sense, a sorting of the columns in the
Cantor fractals of Fig. \ref{fig:cantor}, with the whitespace between columns
removed. For instance, the clustered Cantor plot of the Even process clearly
contains the two major states, with a third ``transient'' state visible
(corresponding to the increasingly unlikely event of never seeing a $0$ in a
block of length $n$). This third state was previously hidden mostly out of view
on the far-right side of the 2-dimensional Cantor plot of the even process in
Fig. \ref{fig:cantor}.

Other features are worth remarking on. Close observation will show the reader
that the hierarchical clustering allows for the (mostly) scale-free distinctions
between pasts with subtle differences. For the $\mathtt{a}^n \mathtt{b}^n$
process, pasts of the form $\dots \mathtt{b a}^n$ are distinguished for
different $n$, as each involves a distinct number of $\mathtt{b}$'s appearing in
the near future; meanwhile, the clustering scheme carefully distinguishes pasts
of the form $\dots \mathtt{b a}^n \mathtt{b}^{n-k}$ for different $k$ but
\emph{not} for different $n$, as $k$ is the essential variable for predicting
the remaining number of $\mathtt{b}$'s. (The scale-free discernment of the
algorithm begins to break down past $n=5$, which is the scale at which sampling
error becomes relevant for our chosen sample size.)

Similar discernment can be noticed for the $\mathtt{a}^n \mathtt{b}^n
\mathtt{c}^n$ and $\mathtt{x+f(x)}$ processes as well. We draw attention to the
manner in which the presence of a semicolon in pasts from $\mathtt{x+f(x)}$
affects the comparison of predictions.

By analyzing clustered Cantor plots, one can gain insight into the properties of
pasts that make them similar in terms of future predictions---even if they are
superficially quite distinct. Furthermore, the horizontal axis allows for
continued use of the natural geometry of the Cantor set for visualizing the
future forecasts associated with each cluster of predictions.

\begin{figure*}[ht]
  \centering
  \includegraphics[width=0.45\linewidth]{../plots/even_MDS.pdf}
  \includegraphics[width=0.45\linewidth]{../plots/anbn_MDS.pdf}
  \caption{Two scatterplots of the first two MDS coordinates of the
  reconstructed predictive states, for the even and $\mathtt{a}^n\mathtt{b}^n$
  processes respectively. The clusters are colored according to the scheme
  determined by the dendrogram in Fig. \ref{fig:hclust} and the label on each
  cluster describes the pattern which uniquely characterizes the pasts in that cluster.}
    \label{fig:mds}
  \Description{Two scatter plots with colored clusters of points are displayed,
  with labels next to each cluster describing symbolic patterns.}
\end{figure*}

\section{Predictive state geometry with multidimensional scaling}
If we are willing to sacrifice direct visualization of future predictions, we
may gain a more intuitive picture of the geometry of predictive state space. We
can apply any desired dimensional reduction algorithm to the matrix of
Wasserstein distances between predictions and obtain a coordinate representation
of the similarities between predictive states.

In Fig. \ref{fig:mds} we have plotted the first two dimensions of a
multidimensional scaling (MDS) decomposition \cite{Borg05a} for the even process
and the $\mathtt{a}^n\mathtt{b}^n$ process, with clusters colored in the same
manner as in Fig. \ref{fig:hclust} and labeled by the specific pattern which
distinguishes the pasts in some of the clusters. It should be noted that the
clusters and the labels are directly drawn from \ref{fig:hclust} for reference,
and are not a result of the MDS algorithm itself. However, interactive plotting
approaches may allow for similar exploration from these decompositions without
the need for prior clustering.

The even process, as in all other cases we have seen thus far, has two dominant
clusters of predictions, corresponding to the predictive states which result
from seeing an even-sized block of $1$'s (or, equivalently, no $1$'s), and that
resulting from seeing an odd-sized block of $1$'s. The
$\mathtt{a}^n\mathtt{b}^n$ plot is much more sophisticated. Intriguingly, its
geometry not only clearly distinguishes predictively distinct states, but it
organizes them in a manner highly suggestive of an \emph{stack}, particularly
appropriate given that stack automata are the natural analogue of hidden Markov
models for context-free processes. As more $\mathtt{a}$'s are observed, we push
further up the stack towards the upper-right corner of the figure, and as more
$\mathtt{b}$'s are observed we pop out of the stack, back towards the lower left
(representing equality between $\mathtt{a}$'s and $\mathtt{b}$'s).

\begin{figure*}[h]
  \centering
  \includegraphics[width=\linewidth]{../plots/feig_MDS.pdf}
  \caption{On the left is a scatterplot of the first two MDS coordinates 
  of the reconstructed predictive states for the Feigenbaum process, 
  color-coded by cluster. On the right, we display a scatterplot of the
  corresponding points in the domain of the logistic map, plotting for each
  point both the present value $y_t$ and the next value $y_{t+1}$, with the
  $x=y$ line for reference. Each pair of color-coded arrows shows where each
  cluster maps to under the action of the logistic map. The predictively
  reconstructed clusters thus correspond to dynamically similar neighborhoods of
  the logistic map domain.}
    \label{fig:mds-feig}
  \Description{A rectangular scatter plot of color-coded points is displayed on
  the left, and on the right a square scatterplot of parabolically arranged
  points with a diagonal line cutting through them. In the right plot the points
  are colored with the same scheme as the left, and similarly colored arrows
  point from clusters to the diagonal line or from the diagonal line to the next
  cluster.}
\end{figure*}

The geometric approach is particularly insightful when we compute the
Wasserstein matrix between predictions estimated from Feigenbaum process data.
Recall that the Feigenbaum process is just a coarse-graining of the iterated
logistic map $y_{t+1} = ry_t(1-y_t)$ at the critical parameter $r\approx
3.56995$. The resulting stream of $0$'s and $1$'s is an infamous instance of
high complexity at the ``boundary of order and chaos'', as values of $r$ on
either side tend to result in coarse-grainings which can be generated by hidden
Markov models, but the Feigenbaum process is context-sensitive and therefores
requires several orders higher of model complexity to capture its behavior.

Regardless, we show that the predictive state geometry reconstructed from a
sufficiently large sample of the Feigenbaum process is capable of recovering the
neighborhoods of $[0,1]$ which are relevant to the dynamics of the original
logistic map. What we mean by this is that there is a correspondence between
each past $x_{-n+1}\dots x_{0}$ and a subset $V_{x_{-n+1}\dots x_{0}}$, such
that $V_{x_1\dots x_n}$ is the set of all points $y$ for which $x(f^{-t}(y)) =
x_{-t}$ for $0\leq t < n$ (here $f(y) = ry(1-y)$ and $x(y)$ is the encoding
$y\mapsto 0,1$). As it happens, pasts $x_{-n+1}\dots x_{0}$ whose predictive
states are close under the Wasserstein distance are also pasts for which the
sets $f(V_{x_{-n+1}\dots x_{0}})$ are close (that is, they correspond to
predictively similar ranges of the logistic map variable).

This relationship between the reconstructed predictive states of the Feigenbaum
process, neighborhoods of the logistic variable $y$, and the logistic map
dynamics is visualized in Fig. \ref{fig:mds-feig}. In short, despite the fact
that the Feigenbaum process is a highly coarse-grained form of the logistic map,
the essential geometry of that map can be recovered by reconstructing predictive
state geometry with the Wasserstein metric and the Cantor embedding.

We should note that, because of the deterministic nature of the Feigenbaum
process, the combination of the Wasserstein metric and the Cantor embedding is
particularly important to achieving this result. Asymptotically, each past
corresponds to a unique future, and so there is no asymptotically no overlap
between predictions. The choice of the Cantor map allows us to place close
together forecasts which match up to a certain time in the future, and the
Wasserstein distance allows us to directly compare predictions whose supports
are geometrically close. It is therefore the combination of these two approaches
which enables the straightforward recovery of the underlying logistic geometry.

\section{Concluding remarks}
We have presented a general approach for predictive state analysis---Cantor
fractal embedding sequences and Wasserstein distance comparison of
predictions---and offered two approaches to visualizing the results of this
method---one a direct application of multidimensional scaling, and the other
being a clustered Cantor diagram built from combining hierarchical clustering
with the introduced Cantor embedding.

In comparison to the use of reproducing kernel Hilbert spaces, which is dominant
approach to predictive states at present
\cite{Song09a,Song10a,Boot13a,Brod20a,Loom21a}, our choice to combine the Cantor
set with the Wasserstein distance may appear rather idiosyncratic, but there are
strong benefits to both methods, and together the two methods synergize their
benefits in a unique way. The topology of convergence in distribution can be
replicated with both the Wasserstein distance and the RKHS inner product, but
the Wasserstein distance depends on far fewer parameters (such as the choice of
the eponymous kernel in RKHS approaches), and its value is directly
interpretable in terms of the shapes of the distributions it compares.

Similarly, there are many ways to metrize the product topology on sequences, but
the Cantor embedding offers a direct way to connect the product topology with a
visualizable geometry, and by embedding in a single dimension enables efficient
computation of the Wasserstein metric. The benefits of the Cantor and
Wasserstein approaches adds interpretability to the resulting predictive state
geometry along two distinct axes, most clearly seen in the clustered Cantor
diagrams of Fig. \ref{fig:hclust}.

%% The acknowledgments section is defined using the "acks" environment % (and
%NOT an unnumbered section). This ensures the proper % identification of the
%section in the article metadata, and the % consistent spelling of the heading.
\begin{acks}
  We thank Nicolas Brodu, Adam Rupe, Alex Jurgens, David Gier, and Mikhael
  Semaan. JPC acknowledges the kind hospitality of the Telluride Science
  Research Center, Santa Fe Institute, Institute for Advanced Study at the
  University of Amsterdam, and California Institute of Technology for their
  hospitality during visits. This material is based upon work supported by, or
  in part by, Grant Nos. FQXi-RFP-IPW-1902 and FQXi-RFP-1809 from the
  Foundational Questions Institute and Fetzer Franklin Fund (a donor-advised
  fund of Silicon Valley Community Foundation) and grants W911NF-18-1-0028 and
  W911NF-21-1-0048 from the U.S. Army Research Laboratory and the U.S. Army
  Research Office.  
\end{acks}

\section{Reproducibility statement}
For the purposes of reproducibility, we have provided a GitHub repository which
contains the code necessary to generate this paper and its figures, including a
notebook for generating the data we used for our examples and building both
static and interactive figures for further exploration.
%%
%% The next two lines define the bibliography style to be used, and % the
%bibliography file.
\bibliographystyle{ACM-Reference-Format}
\bibliography{ref}

%%
%% If your work has an appendix, this is the place to put it.

\end{document}
\endinput
%%
%% End of file `sample-lualatex.tex'.

### 2022-02-06 paper/manuscript.tex
JPC: Done with edit pass. Ready to submit to KDD.

@@
 \begin{abstract}
-  Predictive states for stochastic processes are a non-parametric and
+  Predictive states for stochastic processes are a nonparametric and
   interpretable construct with relevance across a multitude of modeling
   paradigms. Recent progress on the self-supervised reconstruction of predictive
-  states from time-series data has focused on the use of reproducing kernel
+  states from time-series data focused on the use of reproducing kernel
   Hilbert spaces. Here, we examine how Wasserstein distances may be used to
   detect predictive equivalences in symbolic data. We construct a
-  finite-dimensional embedding of sequences based on the Cantor set, and use
+  finite-dimensional embedding of sequences based on the Cantor set and use
   distances in this embedding to compute Wasserstein distances between
   distributions over sequences (``predictions''). We show that analyzing the
-  resulting geometry (via hierarchical clustering and dimension reduction)
+  resulting geometry via hierarchical clustering and dimension reduction
   provides insight into the temporal structure of processes ranging from the
-  relatively simple (\emph{e.g.}\ hidden Markov models) to the very complex
-  (\emph{e.g.}\ indexed grammars).
+  relatively simple (\emph{e.g.}, finite-state hidden Markov models) to the
+  very complex (\emph{e.g.}, infinite-state indexed grammars).
 \end{abstract}
@@
 %being presented. Separate the keywords with commas.
-\keywords{time series, predictive states, wasserstein distance, hierarchical clustering}
+\keywords{time series, predictive states, Cantor set, Wasserstein distance, hierarchical clustering}
 
@@
 observations drawn from a temporal process. We may suppose that the process is
-stationary (time-translation-invariant) and ergodic (has a tendency to explore
+stationary (time-translation invariant) and ergodic (explores
 all possible behaviors) \cite{Kall01a,Shal07a}. We may wish to forecast from the
 observed information the behavior of the next $n$ observations. If we suspect
-that temporal correlations of the process do not matter much beyond $k$ symbols,
-then we can use the data we have to take all subsequences of length $n+k$,
+that the process' temporal correlations do not matter much beyond $k$ symbols,
+then we take our data to be all subsequences of length $n+k$,
 splitting each subsequence into words of length $k$ and $n$ respectively. From
-these ``past/future'' pairs we could construct an empirical conditional
+these ``past/future'' pairs we can construct an empirical conditional
 distribution
-\begin{equation}\label{eq:conditional}
-  \hat{P}_{x_{1}\dots x_{n}|x_{-k+1}\dots x_0} = 
+\begin{align*}
+  \widehat{P}_{x_{1}\dots x_{n}|x_{-k+1}\dots x_0} = 
   \frac{C_{x_{-k+1}\dots x_{n}}}{C_{x_{-k+1}\dots x_0}}
-\end{equation}
+  ~,
+\end{align*}
 where $C_{w}$ is the number of times the word $w$ appears in our sequence $x_1
@@
 
-But suppose now that we do not know how long-range the temporal dependencies in
-our process can stretch. Even very simple stochastic processes can have infinite
-Markov order, indicating potential long-term dependence of future observations
-on the past \cite{Uppe97a}. Given sufficient data, it would be desirable to take
-pasts of arbitrary length, and converge towards some prediction conditioned on
-the \emph{infinite} past:
-\begin{equation}\label{eq:convergence}
+Suppose, though, that we do not know how long-range the process' temporal
+dependencies can stretch. Even very simple stochastic processes can have
+infinite Markov order, indicating potential long-term dependence of future
+observations on the past \cite{Uppe97a}. Given sufficient data, it would be
+desirable to take pasts of arbitrary length and converge towards a prediction
+conditioned on the \emph{infinite} past:
+\begin{align}
   {P}_{x_{1}\dots x_{n}|\overleftarrow{x}}
   = \lim_{k\rightarrow\infty}
-  \hat{P}_{x_{1}\dots x_{n}|x_{-k}\dots x_0}
-\end{equation}
-where $\overleftarrow{x}$ denotes an infinite sequence $\overleftarrow{x} =
-(\dots,x_{-1},x_0)$ of observations stretching into the past. This mathematical
-ideal is known to previous literature as the \emph{causal} or \emph{predictive
-state} \cite{Crut88a,Jaeg00a}. Formally, the conditional predictions
-${P}_{x_{1}\dots x_{n}|\overleftarrow{x}}$ for all forecast lengths $n$ together
-describe a \emph{probability measure} over future sequences $\overrightarrow{x}
-= (x_1,x_2,\dots)$; the predictive state is this measure. We will denote it
-simply by $P_{\overleftarrow{x}}$.
+  \widehat{P}_{x_{1}\dots x_{n}|x_{-k}\dots x_0}
+\label{eq:convergence}
+\end{align}
+with the infinite sequence $\overleftarrow{x} = (\dots,x_{-1},x_0)$ of
+observations stretching into the past. This mathematical ideal is known as the
+\emph{causal} or \emph{predictive state} \cite{Crut88a,Jaeg00a}. Formally, the
+conditional predictions ${P}_{x_{1}\dots x_{n}|\overleftarrow{x}}$ for all
+forecast lengths $n$ together describe a \emph{probability measure} over future
+sequences $\overrightarrow{x} = (x_1,x_2,\dots)$; the predictive state is this
+measure. We denote it simply $P_{\overleftarrow{x}}$.
 
-Predictive states have been utilized for inference and modeling in dynamical
+Predictive states are employed for inference and modeling in dynamical
 systems \cite{Crut92c,Emen16a,Salo19a}, renewal processes and spike-trains
-\cite{Marz14e,Marz15a}, condensed matter physics \cite{Varn12a,Varn14a} and
-spatiotemporal systems \cite{Rupe19a,Rupe19b,Rupe20a}, and a deep mathematical
-theory of predictive state inference has been correspondingly developed
+\cite{Marz14e,Marz15a}, condensed matter physics \cite{Varn12a,Varn14a}, and
+spatiotemporal systems \cite{Rupe19a,Rupe19b,Rupe20a}. A deep mathematical
+theory of predictive-state inference has been correspondingly developed
 \cite{Jaeg00a,Shal02a,Jame04a,Stil07b,Stre13a,Thon15a,Marz19a,Brod20a}. Despite
-this, universal conditions for predictive state convergence were not given until
+this, universal conditions for predictive-state convergence were not given until
 recently. It is now known that if the dataset $x_1 \dots x_L$ is drawn from any
-stationary process, and if $L$ is sufficiently large, then Eq.
-\eqref{eq:convergence} will converge for any $n$ and a probability-1 subset of
+stationary process and if $L$ is sufficiently large, then Eq.
+\eqref{eq:convergence} converges for any $n$ and a probability-$1$ subset of
 pasts $\overleftarrow{x}$ \cite{Loom21a}. In the language of measures,
@@
 
-The general goals of predictive state analysis are typically threefold
-\cite{Shal01a}. The first is to understand the overall structure of how the
-predictive states relate to one another geometrically, and possibly use this
-geometry to classify pasts based on equivalence of their predictive states. The
-second is to actually reproduce the prediction to a specified accuracy. The
-third is to understand the dynamics of how predictions evolve under a stream of
-new observations. We will focus in this paper on the first, as it is a crucial
-building block to achieving the other two.
+The broad goals of predictive-state analysis are threefold \cite{Shal98a}. The
+first is to understand the overall structure of how the predictive states
+relate to one another geometrically and, possibly, use this geometry to
+classify pasts based on equivalence of their predictive states. The second is
+to actually reproduce the prediction to a specified accuracy. The third is to
+understand the dynamics of how predictions evolve under a stream of new
+observations. The following focuses on the first, as it is a crucial building
+block to achieving the other two.
 
-Recent attempts at reconstructing the geometry of predictive states have focused
-on embedding them in reproducing kernel Hilbert spaces
-\cite{Song09a,Song10a,Boot13a,Brod20a,Loom21a}, to great effect largely because
-convergence in Hilbert spaces generated by universal kernels is equivalent to
-convergence in distribution \cite{Srip10a}. That is, these embeddings allow
-accurate representations of predictive states because they respect the product
-topology of sequences in the same way that predictive states themselves do.
+Recent attempts to reconstruct the geometry of predictive states embedded them
+in reproducing kernel Hilbert spaces
+\cite{Song09a,Song10a,Boot13a,Brod20a,Loom21a}. This was achieved to great
+effect largely since convergence in Hilbert spaces generated by universal
+kernels is equivalent to convergence in distribution \cite{Srip10a}. That is,
+these embeddings allow accurate representations of predictive states because
+they respect the product topology of sequences in the same way that predictive
+states themselves do.
 
-Understanding the source of the power of RKHS methods in predictive state
-analysis frees us to consider other options. In this paper, we will embed
+Understanding the source of the power of RKHS methods in predictive-state
+analysis frees us to consider other options. The following embeds
 symbolic sequences in a one-dimensional space that has the same topology as the
@@
 this one-dimensional space. We then use the Wasserstein distance to compute the
-geometry between predictive states, which can be computed with a closed-form
-integral for one-dimensional distributions. This works as an alternative to
-RKHS-based distances because the Wasserstein distance also reproduces the
+geometry between predictive states, which is determined by a closed-form
+integral for one-dimensional distributions. This operates as an alternative to
+RKHS-based distances since the Wasserstein distance also reproduces the
 topology of convergence in distribution \cite{Pane19a}. The resulting distance
-matrix can be used to find low-dimensional embeddings \cite{Borg05a} of the
+matrix then is used to find low-dimensional embeddings \cite{Borg05a} of the
 geometry or hierarchical clusterings \cite{Mull11a} of the predictive states.
-The latter, in particular, when combined with our fractal embedding, provides a
-highly interpretable visualization of the predictive state space.
+When combined with the fractal embedding, the latter, in particular, provides a
+highly interpretable visualization of the predictive-state space.
 
 \section{Example processes}
-The methods in this paper are intended to be applied toward stationary and
-ergodic stochastic processes which produce categorical time-series data. For our
-intents we can think of a stochastic process as a collection of probability
+
+The methods here are intended to be applied to stationary and ergodic
+stochastic processes that generate categorical time-series data. For these
+purposes we consider a stochastic process to be a collection of probability
 distributions $\Prob{\mu}{x_1\dots x_L}$ over any finite, contiguous sequence,
-taking values in a finite set $\mathcal{X}$. Formally this describes a measure
-$\mu$ over the set of all bi-infinite sequences $(\dots,x_{-1},x_0,x_1,\dots)\in
-\mathcal{X}^{\mathbb{Z}}$. These sorts of processes can be generated by a number
-of systems with widely varying complexity. Most popularly studied are those
-methods often characterized as having some degree of ``finite memory'': Markov
-models, hidden Markov models, and observable operator models (also termed
-generalized hidden Markov models) \cite{Uppe97a,Jaeg00a}. Beyond these one can
-also generate processes using probabilistic grammars, such as probabilistic
-context-free and indexed grammars \cite{Gema00a}. Additionally, coarse-grained
-data from dynamical systems (such as the logistic map) can display behavior
-varying widely in complexity \cite{Crut92c}. In this paper we will refer back
-frequently to the following example processes:
+taking values in a finite set $\mathcal{X}$. Formally, this describes a measure
+$\mu$ over the set of all bi-infinite sequences $(\dots,x_{-1},x_0,x_1,\dots)
+\in \mathcal{X}^{\mathbb{Z}}$.
+
+These processes are generated by a number of systems with widely varying
+complexity. Most popularly studied are those often characterized as having a
+degree of ``finite memory'': \emph{Markov chains}, \emph{hidden Markov chains},
+and \emph{observable operator models} (also termed \emph{generalized hidden
+Markov chains}) \cite{Uppe97a,Jaeg00a}. Beyond these, one can also generate
+processes using probabilistic grammars, such as probabilistic context-free and
+indexed grammars \cite{Gema00a}. Additionally, coarse-grained data from chaotic
+dynamical systems---such as the \emph{logistic map}---display behavior varying
+widely in complexity \cite{Crut92c}.
+
+We refer back frequently to the following example processes:
 \begin{enumerate}
 \item The \emph{even process} can be generated by repeatedly tossing a coin
-  and writing down a $0$ for every tail and $11$ for every head. The process is
-  essentially random except for the constraint that $1$'s can only appear in
-  contiguous blocks of even size. The even process has infinite Markov order but
-  can be generated by a two-state hidden Markov process \cite{Crut01a}.
+	and writing down a $0$ for every tail and $11$ for every head. The process
+	is essentially random except that $1$s only appear in contiguous blocks of
+	even size bounded by $0$s. The even process has infinite Markov order but
+	can be generated by a two-state hidden Markov chain \cite{Crut01a}.
 \item The $a^n b^n$ \emph{process} can be generated by choosing a random
-  integer $n\geq 1$ (we will suppose via a Poisson process) and writing $n$
+	integer $n\geq 1$ (we suppose via a Poisson process) and writing $n$
 	$\mathtt{a}$'s followed by an equal number of $\mathtt{b}$'s, and then
-  repeating this process indefinitely. The result will be sequences where any
-  contiguous block of $\mathtt{a}$'s is followed by a block of $\mathtt{b}$'s of
-  equal size. The $\mathtt{a}^n \mathtt{b}^n$ process cannot be modeled by any
-  hidden Markov model, though it is a simple example of a probabilistic
-  context-free grammar \cite{Hopc06a}.
-  \item The $x+f(x)$ \emph{process} is a probabilistic context-free grammar
-  modeling simple mathematical expressions. It has terminal symbols
-  $\{\mathtt{(}\ ,\ \mathtt{)}\ ,\ \mathtt{;}\ ,\ \mathtt{+}\ ,\ \mathtt{f}\ ,\
-  \mathtt{x}\}$ and non-terminals $\{A,B,C\}$, and starts with a sequence osf
-  $A$'s with the production rules:
+	repeating this procedure indefinitely. This results in sequences where any
+	contiguous block of $\mathtt{a}$s is followed by a block of $\mathtt{b}$s
+	of equal size. The $\mathtt{a}^n \mathtt{b}^n$ process cannot be generated
+	by any finite hidden Markov chain, though it is a simple example of a
+	probabilistic context-free language \cite{Hopc06a}.
+\item The $x+f(x)$ \emph{process} is a probabilistic context-free language
+	modeling the syntactic structure of simple mathematical expressions. It has
+	terminal symbols $\{\mathtt{(}\ ,\ \mathtt{)}\ ,\ \mathtt{;}\ ,\
+	\mathtt{+}\ ,\ \mathtt{f}\ ,\ \mathtt{x}\}$ and nonterminals $\{A,B,C\}$,
+	and starts with a sequence of $A$s. Sequences are generated by applying the
+	production rules:
 \begin{align*}
@@
     C &\mapsto \mathtt{f(}B\mathtt{)}\ |\ \mathtt{x}
+  ~.
 \end{align*}
-  \item The $a^n b^n c^n$ \emph{process} is a probabilistic indexed grammar
-  \cite{Hopc06a} which is analagous to $\mathtt{a}^n \mathtt{b}^n$ except after
-  writing the blocks of $\mathtt{a}$'s and $\mathtt{b}$'s, we also write a block
-  of $\mathtt{c}$'s of length $n$.
-  \item The \emph{Feigenbaum process} is generated by sampling from the time
-  series of the logistic map at critical parameter $r\approx 3.56995$:
-  \begin{equation*}
+\item The $a^n b^n c^n$ \emph{process} is a probabilistic indexed language
+	\cite{Hopc06a} that is analogous to $\mathtt{a}^n \mathtt{b}^n$ except
+	after writing the blocks of $\mathtt{a}$'s and $\mathtt{b}$'s, we also
+	write a block of $\mathtt{c}$'s of length $n$.
+\item The \emph{Morse-Thue process} is generated by sampling from the time
+	series of the logistic map at critical ``onset of chaos'' parameter
+	$r_c \approx 3.56995$:
+\begin{align*}
     y_{t+1} = r y_t(1-y_t)
-  \end{equation*}
+\end{align*}
 	and then coarse-graining the data by taking $x_t = 0$ if $0< y_t \leq
 	\frac{1}{2}$ and $x_t = 1$ if $\frac{1}{2} < y_t <1$ \cite{Kurk03a}.
-  Alternatively, we can generate this process by starting with a single $0$ and
-  executing the replacements $0\mapsto 11$ and $1 \mapsto 01$ consecutively. The
-  result of the Feigenbaum process is indexed-context free \cite{Crut92c}.
+	Alternatively, we can generate this process by starting with a single $0$
+	and executing the replacements $0 \mapsto 11$ and $1 \mapsto 01$
+	consecutively. The resulting process is an indexed-context free language
+	\cite{Crut92c}.
 \end{enumerate}
 
-\section{Cantor-embedding sequences}
 \begin{figure*}[h]
@@
 \includegraphics[width=0.4\linewidth]{../plots/x+f(x)_CANTOR.png}
-  \caption{ Cantor plots for the even, $\mathtt{a}^n\mathtt{b}^n$,
+\caption{Cantor plots for the even, $\mathtt{a}^n\mathtt{b}^n$,
 	$\mathtt{a}^n\mathtt{b}^n\mathtt{c}^n$, and $\mathtt{x+f(x)}$ processes.
 	Each point $(x,y)$ corresponds to a pair of sequences corresponding to the
-    past and future respectively. The symbol on the $x$ ($y$) axis indicate that
-    all points above (to the right of) that symbol have a past (future) whose
-    most recent observation is that symbol. Though not marked, further
-    proportional subdivisions of each segment of the axes would indicate the
-    value of the second, third, \emph{etc.} symbols. For instance, one can read
-    from the $\mathtt{x+f(x)}$ fractal that any past which ends in $\mathtt{f}$
-    must be paired with a future beginning in $\mathtt{(f}$ or $\mathtt{(x}$. }
+	past and future, respectively. The symbol on the $x$ ($y$) axis indicates
+	that all points above (to the right of) that symbol have a past (future)
+	whose most recent observation is that symbol. Though not marked, further
+	proportional subdivisions of each segment of the axes indicate the value of
+	the second, third, \emph{etc.} symbols. For instance, one can read from the
+	$\mathtt{x+f(x)}$ fractal that any past ending in $\mathtt{f}$ must be
+	paired with a future beginning in $\mathtt{(f}$ or $\mathtt{(x}$.
+	}
 \label{fig:cantor}
@@
 	$\mathtt{a}^n\mathtt{b}^n$, $\mathtt{a}^n\mathtt{b}^n\mathtt{c}^n$, and
-  $\mathtt{x+f(x)}$ processes.}
+	$\mathtt{x+f(x)}$ processes.
+	}
 \end{figure*}
+
+\section{Cantor-embedding sequences}
 
@@
 
-We can exploit this self-similarity in an interesting way by constructing a
-mapping between sequence space and the celebrated Cantor set (or one of its
+We exploit this self-similarity in an interesting way by constructing a mapping
+between sequence space and the celebrated \emph{Cantor set} (or one of its
 generalizations). Suppose a symbolic sequence $(x_1,x_2,\dots)$ takes values in
 an alphabet $\mathcal{X}$ of size $|\mathcal{X}|$. To each $x\in\mathcal{X}$ we
-can associate some unique integer between $0$ and $|\mathcal{X}|-1$ inclusive;
-call this $J(x)$. Then there is a function $C:\mathcal{X}^\mathbb{N}\rightarrow
-[0,1]$ which maps every sequence to a positive real number:
-\begin{equation}
+associate a unique integer between $0$ and $|\mathcal{X}|-1$ inclusive; call
+this $J(x)$. Then, there is a function $C:\mathcal{X}^\mathbb{N}\rightarrow
+[0,1]$ that maps every sequence to a positive real number:
+\begin{align*}
   C(x_1,x_2,\dots) = \sum_{k=1}^\infty \frac{2J(x_k)}{(2|\mathcal{X}|-1)^k}
-\end{equation}
+  ~.
+\end{align*}
 For instance, suppose that $|\mathcal{X}|=2$ has two elements; then $C$ maps the
@@
 $f$ on $[0,1]$, the function $F(\overrightarrow{x}) = f(C(\overrightarrow{x}))$
-is continuous on $\mathcal{X}^\mathbb{N}$; further, if $F$ is continuous on
+is continuous on $\mathcal{X}^\mathbb{N}$. Further, if $F$ is continuous on
 $\mathcal{X}^\mathbb{N}$, then $f(y) = F(C^{-1}(y))$ is continuous on the image.
-Thus the embedding $C$ respects the basic structure of the product topology
+Thus, the embedding $C$ respects the basic structure of the product topology
 \cite{Kurk03a}.
@@
 Stationary processes, due to their time-translation invariance, inherit the
-fractal temporality of sequence space. This can be neatly visualized: given a
-length-$L$ sample $x_1\dots x_L$, and some $n,k>0$, take a sliding window of
-pasts and futures, $(x_{t-n+1}\dots x_t, x_{t+1}\dots x_k)$ for $t=n,\dots,L-k$.
-For each past-future pair, compute the truncated Cantor embeddings on the
+fractal temporality of sequence space. This can be easily visualized: Given a
+length-$L$ sample $x_1\dots x_L$, and $n,k>0$, take a sliding window of pasts
+and futures, $(x_{t-n+1}\dots x_t, x_{t+1}\dots x_k)$ for $t=n,\dots,L-k$.  For
+each past-future pair, compute the truncated Cantor embeddings on the
 \emph{reversed} past and (unreversed) future: $\left(C(x_t \dots
 x_{t-n+1}),C(x_{t+1}\dots x_k)\right)$. The resulting pairs of real numbers can
-be plotted as $(x,y)$-values on a scatter plot. The fractal which emerges
-contains, in essence, all information necessary to understand the temporal
-structures of the process. See Fig. \ref{fig:cantor} for examples and guidance
-on how to interpret the visualization.
+be plotted as $(x,y)$-values on a scatter plot. The fractal that emerges
+contains, in essence, all information necessary to understand a process'
+temporal structures. See Fig. \ref{fig:cantor} for examples and guidance on how
+to interpret the visualization.
 
-It is worth noting that for $|\mathcal{X}|>2$, the embedding $C$ also introduces
-additional structure which may or may not be desirable. The association of each
-symbol $x$ with an integer $j_x$ endows an ordinal structure on the set
-$\mathcal{X}$; this ordinality is present in the macroscopic geometry of
-$C\left(\mathcal{X}^{\mathbb{N}}\right)$. Later we will examine higher-dimension
-embeddings which do not have this ordinal aspect; however, they will come at the
-cost of increased computational complexity for the Wasserstein distance
-computation, so for now we will assume that ordinal artifacts are either desired
-or sufficiently tolerable as to not outweight the computational benefits of
-working in one dimension.
-
-\section{Wasserstein distance on predictive states}
+Note that for $|\mathcal{X}|>2$ the embedding $C$ introduces additional
+structure that may or may not be desired. Associating each symbol $x$ with an
+integer $j_x$ endows an ordinal structure on the set $\mathcal{X}$. This
+ordinality is present in the macroscopic geometry of
+$C\left(\mathcal{X}^{\mathbb{N}}\right)$. Later, we examine higher-dimension
+embeddings that do not have this. However, they come at the cost of increased
+computational complexity when determining the Wasserstein distance. So, for now,
+we assume that ordinal artifacts are either desired or sufficiently tolerable
+as to not outweigh the computational benefits of working in one dimension.
 
@@
 
-The Cantor fractals in Fig. \ref{fig:cantor} can be thought of as probability
-distributions. We can consequently interpret a vertical slices of the fractal,
-located at horizontal position $C(\overleftarrow{x})$, as visualizing the
-predictive state $P_{\overleftarrow{x}}$ as a distribution over Cantor-embedded
-futures, $C(\overrightarrow{x})$.
+\section{Wasserstein distance on predictive states}
 
-For example, by analyzing the Cantor fractal of the even process, one may notice
-that there are effectively only 2 distinct predictive states---every vertical
-column is just one of two types. This corresponds with the 2 states of the
-hidden Markov model which generates the even process.
+Figure \ref{fig:cantor}'s Cantor fractals represent probability distributions:
+We interpret a vertical slice of the fractal, located at horizontal position
+$C(\overleftarrow{x})$, as visualizing the predictive state
+$P_{\overleftarrow{x}}$ as a distribution over Cantor-embedded futures
+$C(\overrightarrow{x})$.
 
-This visualization allows us to see how predictive states distribute their
-probability over the intrinsic geometry of potential futures. We can compare
-predictive states not only on how much their supports overlap, but on how
-geometrically close their supports are to one another. For the $\mathtt{a}^n
-\mathtt{b}^n$ process, for example, we can see that the first few columns
-(corresponding to pasts of the form $\dots \mathtt{b a}^n$ for some $n$) are
-inherently similar to one another, though they are shifted upwards the closer to
-the axis they are (which corresponds to the increasing number of $\mathtt{b}$'s
-in the predicted future as $n$ increases).
+For example, by examining the even process' Cantor fractal, one notices that
+there are effectively only $2$ distinct predictive states---every vertical
+column is just one of two types. This corresponds with the $2$ states of the
+hidden Markov model that generates the even process.
+
+This allows us to see how predictive states distribute their probability over
+the intrinsic geometry of potential futures. We compare predictive states not
+only on how much their supports overlap, but on how geometrically close their
+supports are to one another. For the $\mathtt{a}^n \mathtt{b}^n$ process, for
+example, we see that the first few columns (corresponding to pasts of the form
+$\dots \mathtt{b a}^n$ for some $n$) are inherently similar to one another,
+though they are shifted upwards the closer to the axis they are. (The latter
+corresponds to the increasing number of $\mathtt{b}$s in the predicted future
+as $n$ increases.)
 
@@
 underlying geometry is the Wasserstein metric \cite{Pane19a}. Given two measures
-$\mu$, $\nu$ defined on a metric space $\mathcal{M}$ with metric $d$, the
-Wasserstein distance between $\mu$ and $\nu$ is given by
-\begin{equation*}
+$\mu$ and $\nu$ defined on a metric space $\mathcal{M}$ with metric $d$, the
+Wasserstein distance between $\mu$ and $\nu$ is given by:
+\begin{align*}
   W(\mu,\nu) = \min_{\pi\in \Gamma(\mu,\nu)}
@@
    d(x,y) d\pi(x,y)
-\end{equation*}
-where $\Gamma(\mu,\nu)$ is the set of all measures on
-$\mathcal{M}\times\mathcal{M}$ whose left and right marginals are $\mu$ and
-$\nu$ respectively. This can be interpreted as the minimal cost for ``shifting''
-the probability mass from one distribution to match shape of the other.
+   ~,
+\end{align*}
+where $\Gamma(\mu,\nu)$ is the set of all measures on $\mathcal{M} \times
+\mathcal{M}$ whose left and right marginals are $\mu$ and $\nu$, respectively.
+It is the minimal cost to ``shift'' the probability mass from one distribution
+to match shape of the other.
 
-$W(\mu,\nu)$ the solution to a constrained linear optimization. As a function of
-distributions, $W(\mu,\nu)$ is continuous with respect to convergence in
-distribution; in fact, convergence under the Wasserstein distance is equivalent
-to convergence in distribution on compact spaces. This makes $W(\mu,\nu)$ ideal
-for measuring geometry between predictive states, since empirical estimates of
-these are known to converge in distribution.
+$W(\mu,\nu)$ is the solution to a constrained linear optimization. As a
+function of distributions, $W(\mu,\nu)$ is continuous with respect to
+convergence in distribution. In fact, convergence under the Wasserstein
+distance is equivalent to convergence in distribution on compact spaces. This
+makes $W(\mu,\nu)$ ideal for measuring geometry between predictive states,
+since empirical estimates of these are known to converge in distribution.
 
@@
 to the Wasserstein optimization problem \cite{Thas10a}. Let $F$ and $G$
-respectively be the cumulative distributions functions of $\mu$ and $\nu$. Then
-\begin{equation}
+respectively be the cumulative distributions functions of $\mu$ and $\nu$.
+Then:
+\begin{align*}
   W(\mu,\nu) = \int_{-\infty}^\infty \left|F(t)-G(t)\right| dt
-\end{equation}
+  ~.
+\end{align*}
 This closed-form solution is considerably faster to compute than the linear
-optimization required for arbitrary metric spaces. Because the Cantor embedding
-embeds the space of sequences directly into $[0,1]$, we can utilize this
-formula.
+optimization required for arbitrary metric spaces. Since the Cantor embedding
+embeds the space of sequences directly into $[0,1]$, we can directly employ
+this formula.
 
-The combination of the Cantor embedding and the Wasserstein distance lays out a
-fairly straightforward programme for analyizing a categorical time series:
+Combining the Cantor embedding and Wasserstein distance leads to a
+straightforward program for analyzing categorical time series:
 \begin{enumerate}
 \item Apply Algorithm \ref{alg:cantorwass} to the data stream $x_1\dots x_L$
-  for a specified past length $k$ and future length $n$, retrieving the set of
-  unique observed pasts, the empirical Cantor distributions corresponding to
-  each past, and the matrix of Wasserstein distances computed from these
-  distributions.
-  \item Use the Wasserstein distance matrix to perform additional methods of
-  geometric data analysis, such as hierarchical clustering \cite{Mull11a} and
-  multidimensional scaling \cite{Borg05a}, in order to elucidate the relative
-  geometry of the predictive states.
+	for a specified past length $k$ and future length $n$, retrieving the (i)
+	set of unique observed pasts, (ii) empirical Cantor distributions
+	corresponding to each past, and (iii) matrix of Wasserstein distances
+	computed from these distributions.
+\item To elucidate the relative geometry of the predictive states, use the
+	Wasserstein distance matrix to perform additional methods of geometric data
+	analysis, such as hierarchical clustering \cite{Mull11a} and
+	multidimensional scaling \cite{Borg05a}.
 \end{enumerate}
-In the next two sections we examine the results of this approach.
+The next two sections examine the results of this approach.
 
-\section{Interpretable predictive states with hierarchical clustering}
 \begin{figure*}[ht]
@@
 \includegraphics[width=0.45\linewidth]{../plots/x+f(x)_HCLUST.png}
-  \caption{From upper left to lower right, the clustered Cantor diagrams of the even, $\mathtt{a}^n\mathtt{b}^n$,
-  $\mathtt{a}^n\mathtt{b}^n\mathtt{c}^n$, and $\mathtt{x+f(x)}$ processes (zoom
-  for detail). On
-  each diagram, the vertical axis shows all pasts of a given length (length
-  $k=8$ for the even, $\mathtt{a}^n\mathtt{b}^n$ and
+\caption{(Upper left to lower right) Clustered Cantor diagrams of the even,
+	$\mathtt{a}^n\mathtt{b}^n$, $\mathtt{a}^n\mathtt{b}^n\mathtt{c}^n$, and
+	$\mathtt{x+f(x)}$ processes. Zoom for detail. For each, the vertical axis
+	shows all pasts of a given length $k$ along with their hierarchically
+	clustered dendrogram. $k=8$ for the even, $\mathtt{a}^n\mathtt{b}^n$, and
 	$\mathtt{a}^n\mathtt{b}^n\mathtt{c}^n$ processes and $k=4$ for the
-  $\mathtt{x+f(x)}$ process), along with their hierarchically clustered
-  dendrogram. For the purposes of this graphic the coloring threshold was
-  chosen to aid in visual interpretation. The lines in each row show the
+	$\mathtt{x+f(x)}$ process. For present purposes, the coloring threshold
+	was chosen to aid visual interpretation. The lines in each row show the
 	empirical distribution of Cantor-embedded futures observed following each
-  past. As such the horizontal axis corresponds exactly to the vertical axis of
-  Fig. \ref{fig:cantor}.}
+	past. As such, the horizontal axis corresponds exactly to the vertical axis
+	of Fig. \ref{fig:cantor}.
+	}
 \label{fig:hclust}
-  \Description{Four graphs are displayed. Each vertical axis has a list of symbolic sequences
-  organized by a hierarchical clustering tree. The horizontal axis is a list of
-  individual symbols. The content of each graph is rows of fractally-distributed
-  lines.}
+\Description{Four graphs are displayed. Each vertical axis has a list of
+	symbolic sequences organized by a hierarchical clustering tree. The
+	horizontal axis is a list of individual symbols. The content of each graph
+	is rows of fractal-distributed lines.
+	}
 \end{figure*}
-In Fig. \ref{fig:hclust} we display the result of collecting the Cantor-embedded
-empirical predictions for all pasts of a given length, for four processes (even,
-$\mathtt{a}^n \mathtt{b}^n$, $\mathtt{a}^n \mathtt{b}^n \mathtt{c}^n$ and
-$\mathtt{x+f(x)}$). In each case the Wasserstein distance between every pair of
-predictions was computed and used to hierarchically cluster the pasts (using the
-Ward method \cite{Mull11a}) with
-others that produced similar predictions.
+
+\section{Interpretable predictive states with hierarchical clustering}
+
+Figure \ref{fig:hclust} displays the result of collecting the Cantor-embedded
+empirical predictions for all pasts of a given length for four processes---even,
+$\mathtt{a}^n \mathtt{b}^n$, $\mathtt{a}^n \mathtt{b}^n \mathtt{c}^n$, and
+$\mathtt{x+f(x)}$. For each, the Wasserstein distance between every pair of
+predictions was computed and used to hierarchically cluster the pasts with
+others that produced similar predictions, using the Ward method \cite{Mull11a}.
 
 The resulting clustered Cantor plots offer a highly interpretable visualization
-of the relationship between pasts and futures, and of the geometry of the
-predictive states. They are, in a certain sense, a sorting of the columns in the
-Cantor fractals of Fig. \ref{fig:cantor}, with the whitespace between columns
-removed. For instance, the clustered Cantor plot of the Even process clearly
-contains the two major states, with a third ``transient'' state visible
-(corresponding to the increasingly unlikely event of never seeing a $0$ in a
-block of length $n$). This third state was previously hidden mostly out of view
-on the far-right side of the 2-dimensional Cantor plot of the even process in
-Fig. \ref{fig:cantor}.
+of the relationship between pasts and futures, and of the predictive states'
+geometry. Each plot, in a certain sense, sorts the columns in the Cantor
+fractals of Fig. \ref{fig:cantor} with the white space between columns removed.
+For instance, the even process's clustered Cantor plot clearly contains the two
+major states, with a third ``transient'' state visible. (The latter corresponds
+to the increasingly unlikely event of never seeing a $0$ in a block of length
+$n$.) This third state was previously hidden mostly out of view on the
+far-right side of the $2$-dimensional Cantor plot of the even process in Fig.
+\ref{fig:cantor}.
 
-Other features are worth remarking on. Close observation will show the reader
-that the hierarchical clustering allows for the (mostly) scale-free distinctions
-between pasts with subtle differences. For the $\mathtt{a}^n \mathtt{b}^n$
-process, pasts of the form $\dots \mathtt{b a}^n$ are distinguished for
-different $n$, as each involves a distinct number of $\mathtt{b}$'s appearing in
-the near future; meanwhile, the clustering scheme carefully distinguishes pasts
-of the form $\dots \mathtt{b a}^n \mathtt{b}^{n-k}$ for different $k$ but
-\emph{not} for different $n$, as $k$ is the essential variable for predicting
-the remaining number of $\mathtt{b}$'s. (The scale-free discernment of the
-algorithm begins to break down past $n=5$, which is the scale at which sampling
-error becomes relevant for our chosen sample size.)
+Other features are worth calling out. Close observation shows that hierarchical
+clustering reveals the (mostly) scale-free distinctions between pasts with
+subtle differences. For the $\mathtt{a}^n \mathtt{b}^n$ process, pasts of the
+form $\dots \mathtt{b a}^n$ are distinguished for different $n$, as each
+involves a distinct number of $\mathtt{b}$'s appearing in the near future.
+Meanwhile, the clustering scheme carefully distinguishes pasts of the form
+$\dots \mathtt{b a}^n \mathtt{b}^{n-k}$ for different $k$ but \emph{not} for
+different $n$, as $k$ is the essential variable for predicting the remaining
+number of $\mathtt{b}$'s. (The scale-free discernment of the algorithm breaks
+down past $n=5$---the scale at which sampling error becomes relevant for our
+chosen sample size.)
 
-Similar discernment can be noticed for the $\mathtt{a}^n \mathtt{b}^n
-\mathtt{c}^n$ and $\mathtt{x+f(x)}$ processes as well. We draw attention to the
-manner in which the presence of a semicolon in pasts from $\mathtt{x+f(x)}$
-affects the comparison of predictions.
+Similar discernment is seen for the $\mathtt{a}^n \mathtt{b}^n \mathtt{c}^n$
+and $\mathtt{x+f(x)}$ processes as well. We draw attention to the manner in
+which the presence of a semicolon in pasts from $\mathtt{x+f(x)}$ affects the
+comparison of predictions.
 
-By analyzing clustered Cantor plots, one can gain insight into the properties of
-pasts that make them similar in terms of future predictions---even if they are
+By analyzing clustered Cantor plots, one gains insight into the properties of
+pasts that make them similar in terms of future predictions, even if they are
 superficially quite distinct. Furthermore, the horizontal axis allows for
-continued use of the natural geometry of the Cantor set for visualizing the
-future forecasts associated with each cluster of predictions.
+continued use of the Cantor set's natural geometry for visualizing the future
+forecasts associated with each cluster of predictions.
 
@@
 \includegraphics[width=0.45\linewidth]{../plots/anbn_MDS.pdf}
-  \caption{Two scatterplots of the first two MDS coordinates of the
-  reconstructed predictive states, for the even and $\mathtt{a}^n\mathtt{b}^n$
-  processes respectively. The clusters are colored according to the scheme
-  determined by the dendrogram in Fig. \ref{fig:hclust} and the label on each
-  cluster describes the pattern which uniquely characterizes the pasts in that cluster.}
+\caption{Scatterplots of the first two MDS coordinates of the reconstructed
+	predictive states: (Left) Even process. (Right) $\mathtt{a}^n\mathtt{b}^n$
+	process. Clusters colored according to the scheme determined by the
+	dendrogram in Fig. \ref{fig:hclust} and the label on each cluster describes
+	the pattern that uniquely characterizes the pasts in that cluster.
+	}
 \label{fig:mds}
 \Description{Two scatter plots with colored clusters of points are displayed,
-  with labels next to each cluster describing symbolic patterns.}
+	with labels next to each cluster describing symbolic patterns.
+	}
 \end{figure*}
@@
 \section{Predictive state geometry with multidimensional scaling}
-If we are willing to sacrifice direct visualization of future predictions, we
-may gain a more intuitive picture of the geometry of predictive state space. We
-can apply any desired dimensional reduction algorithm to the matrix of
-Wasserstein distances between predictions and obtain a coordinate representation
-of the similarities between predictive states.
 
-In Fig. \ref{fig:mds} we have plotted the first two dimensions of a
-multidimensional scaling (MDS) decomposition \cite{Borg05a} for the even process
-and the $\mathtt{a}^n\mathtt{b}^n$ process, with clusters colored in the same
-manner as in Fig. \ref{fig:hclust} and labeled by the specific pattern which
-distinguishes the pasts in some of the clusters. It should be noted that the
-clusters and the labels are directly drawn from \ref{fig:hclust} for reference,
-and are not a result of the MDS algorithm itself. However, interactive plotting
+Sacrificing direct visualization of future predictions leads to a more
+intuitive picture of predictive-state space geometry. Applying any desired
+dimension reduction algorithm to the matrix of Wasserstein distances between
+predictions yields a coordinate representation of the similarities between
+predictive states.
+
+Figure \ref{fig:mds} plots the first two dimensions of a multidimensional
+scaling (MDS) decomposition \cite{Borg05a} for the even and
+$\mathtt{a}^n\mathtt{b}^n$ processes. Clusters are colored in the same manner
+as in Fig. \ref{fig:hclust} and labeled by the specific pattern that
+distinguishes the pasts in some of the clusters. Note that the clusters and
+labels are directly drawn from Fig. \ref{fig:hclust} for reference. They are
+not the result of the MDS algorithm itself. However, interactive plotting
 approaches may allow for similar exploration from these decompositions without
@@
 
-The even process, as in all other cases we have seen thus far, has two dominant
-clusters of predictions, corresponding to the predictive states which result
-from seeing an even-sized block of $1$'s (or, equivalently, no $1$'s), and that
-resulting from seeing an odd-sized block of $1$'s. The
-$\mathtt{a}^n\mathtt{b}^n$ plot is much more sophisticated. Intriguingly, its
-geometry not only clearly distinguishes predictively distinct states, but it
-organizes them in a manner highly suggestive of an \emph{stack}, particularly
-appropriate given that stack automata are the natural analogue of hidden Markov
-models for context-free processes. As more $\mathtt{a}$'s are observed, we push
-further up the stack towards the upper-right corner of the figure, and as more
-$\mathtt{b}$'s are observed we pop out of the stack, back towards the lower left
-(representing equality between $\mathtt{a}$'s and $\mathtt{b}$'s).
+The even process, as in all other cases seen thus far, has two dominant
+prediction clusters. These correspond to the predictive states that result
+from seeing an even-sized block of $1$s (or, equivalently, no $1$s), and that
+result from seeing an odd-sized block of $1$s.
+
+The $\mathtt{a}^n\mathtt{b}^n$ plot is much more sophisticated. Intriguingly,
+its geometry not only clearly distinguishes predictively distinct states, but
+organizes them in a manner highly suggestive of an \emph{pushdown stack}. The
+latter is particularly appropriate given that stack automata are the natural
+analog of hidden Markov chains but for context-free languages. Observing more
+$\mathtt{a}$s pushes more symbols onto the stack, with the predictive states
+moving further up towards the plot's upper-right corner. And, as more
+$\mathtt{b}$'s are observed the top symbol is popped off the stack, and the
+predictive states move back towards the lower left. The latter represents
+equality between $\mathtt{a}$s and $\mathtt{b}$s.
 
@@
 \includegraphics[width=\linewidth]{../plots/feig_MDS.pdf}
-  \caption{On the left is a scatterplot of the first two MDS coordinates 
-  of the reconstructed predictive states for the Feigenbaum process, 
-  color-coded by cluster. On the right, we display a scatterplot of the
-  corresponding points in the domain of the logistic map, plotting for each
-  point both the present value $y_t$ and the next value $y_{t+1}$, with the
-  $x=y$ line for reference. Each pair of color-coded arrows shows where each
-  cluster maps to under the action of the logistic map. The predictively
-  reconstructed clusters thus correspond to dynamically similar neighborhoods of
-  the logistic map domain.}
+\caption{(Left) Scatterplot of the first two MDS coordinates of the
+	reconstructed predictive states for the Morse-Thue process, color-coded by
+	cluster. (Right) Scatterplot of the corresponding points in the domain of
+	the logistic map, plotting for each point both the present value $y_t$ and
+	the next value $y_{t+1}$, with the $x=y$ line for reference. Each pair of
+	color-coded arrows shows where each cluster maps to under the action of the
+	logistic map. The predictively reconstructed clusters thus correspond to
+	dynamically similar neighborhoods of the logistic map domain.
+	}
 \label{fig:mds-feig}
@@
 	the left, and on the right a square scatterplot of parabolically arranged
-  points with a diagonal line cutting through them. In the right plot the points
-  are colored with the same scheme as the left, and similarly colored arrows
-  point from clusters to the diagonal line or from the diagonal line to the next
-  cluster.}
+	points with a diagonal line cutting through them. In the right plot the
+	points are colored with the same scheme as the left, and similarly colored
+	arrows point from clusters to the diagonal line or from the diagonal line
+	to the next cluster.
+	}
 \end{figure*}
 
-The geometric approach is particularly insightful when we compute the
-Wasserstein matrix between predictions estimated from Feigenbaum process data.
-Recall that the Feigenbaum process is just a coarse-graining of the iterated
-logistic map $y_{t+1} = ry_t(1-y_t)$ at the critical parameter $r\approx
-3.56995$. The resulting stream of $0$'s and $1$'s is an infamous instance of
-high complexity at the ``boundary of order and chaos'', as values of $r$ on
-either side tend to result in coarse-grainings which can be generated by hidden
-Markov models, but the Feigenbaum process is context-sensitive and therefores
-requires several orders higher of model complexity to capture its behavior.
+The geometric approach is particularly insightful when computing the
+Wasserstein matrix between predictions estimated from Morse-Thue process data.
+Recall that the Morse-Thue process is a coarse-graining of the iterated
+logistic map $y_{t+1} = ry_t(1-y_t)$ at the critical chaos parameter $r_c
+\approx 3.56995$. The resulting stream of $0$s and $1$s is a well-known
+instance of high complexity at the ``order-disorder border''. Specifically,
+setting parameter $r$ on either side of $r_c$ results in sequences that
+can be generated by finite hidden Markov chains. However, at $r_c$ itself the
+resulting Morse-Thue process is context-sensitive and therefore requires
+infinite predictive states. That is, when it comes to capturing its behavior,
+the process is several orders higher in model complexity. It is further up the
+Chomsky language hierarchy.
 
-Regardless, we show that the predictive state geometry reconstructed from a
-sufficiently large sample of the Feigenbaum process is capable of recovering the
-neighborhoods of $[0,1]$ which are relevant to the dynamics of the original
-logistic map. What we mean by this is that there is a correspondence between
-each past $x_{-n+1}\dots x_{0}$ and a subset $V_{x_{-n+1}\dots x_{0}}$, such
-that $V_{x_1\dots x_n}$ is the set of all points $y$ for which $x(f^{-t}(y)) =
-x_{-t}$ for $0\leq t < n$ (here $f(y) = ry(1-y)$ and $x(y)$ is the encoding
-$y\mapsto 0,1$). As it happens, pasts $x_{-n+1}\dots x_{0}$ whose predictive
+Despite this high order of structural complexity, the predictive state geometry
+reconstructed from a sufficiently large sample of the Morse-Thue process
+recovers the neighborhoods of $[0,1]$ that are relevant to the dynamics of the
+original logistic map. Said differently, there is a correspondence between each
+past $x_{-n+1}\dots x_{0}$ and a subset $V_{x_{-n+1}\dots x_{0}}$, such that
+$V_{x_1\dots x_n}$ is the set of all points $y$ for which $x(f^{-t}(y)) =
+x_{-t}$ for $0\leq t < n$. (Here, $f(y) = ry(1-y)$ and $x(y)$ is the encoding
+$y\mapsto \{0,1\}$.) As it happens, pasts $x_{-n+1}\dots x_{0}$ whose predictive
 states are close under the Wasserstein distance are also pasts for which the
-sets $f(V_{x_{-n+1}\dots x_{0}})$ are close (that is, they correspond to
-predictively similar ranges of the logistic map variable).
+sets $f(V_{x_{-n+1}\dots x_{0}})$ are close. That is, they correspond to
+predictively similar ranges of the logistic map variable.
 
-This relationship between the reconstructed predictive states of the Feigenbaum
-process, neighborhoods of the logistic variable $y$, and the logistic map
-dynamics is visualized in Fig. \ref{fig:mds-feig}. In short, despite the fact
-that the Feigenbaum process is a highly coarse-grained form of the logistic map,
-the essential geometry of that map can be recovered by reconstructing predictive
-state geometry with the Wasserstein metric and the Cantor embedding.
+Figure \ref{fig:mds-feig} directly visualizes the relationship between the
+reconstructed predictive states of the Morse-Thue process, neighborhoods of the
+logistic variable $y$, and the logistic map dynamics. In short, despite the
+fact that the Morse-Thue process is a highly coarse-grained form of the
+logistic map, the essential geometry of that map can be recovered by
+reconstructing predictive state geometry with the Wasserstein metric and the
+Cantor embedding.
 
-We should note that, because of the deterministic nature of the Feigenbaum
-process, the combination of the Wasserstein metric and the Cantor embedding is
-particularly important to achieving this result. Asymptotically, each past
-corresponds to a unique future, and so there is no asymptotically no overlap
-between predictions. The choice of the Cantor map allows us to place close
-together forecasts which match up to a certain time in the future, and the
-Wasserstein distance allows us to directly compare predictions whose supports
-are geometrically close. It is therefore the combination of these two approaches
-which enables the straightforward recovery of the underlying logistic geometry.
+Note that, due to the deterministic nature of the Morse-Thue process, the
+combination of the Wasserstein metric and the Cantor embedding is particularly
+important to achieving this result. Asymptotically, each past corresponds to a
+unique future. And so, there is asymptotically no overlap between predictions.
+The choice of the Cantor map facilitates placing together forecasts that match
+up to a certain time in the future. And, the Wasserstein distance allows
+directly comparing predictions whose supports are geometrically close. In this
+way, the combination of the two approaches enables the straightforward recovery
+of the underlying dynamical system's (logistic map's) geometry.
 
 \section{Concluding remarks}
-We have presented a general approach for predictive state analysis---Cantor
-fractal embedding sequences and Wasserstein distance comparison of
-predictions---and offered two approaches to visualizing the results of this
-method---one a direct application of multidimensional scaling, and the other
-being a clustered Cantor diagram built from combining hierarchical clustering
-with the introduced Cantor embedding.
 
-In comparison to the use of reproducing kernel Hilbert spaces, which is dominant
-approach to predictive states at present
-\cite{Song09a,Song10a,Boot13a,Brod20a,Loom21a}, our choice to combine the Cantor
-set with the Wasserstein distance may appear rather idiosyncratic, but there are
-strong benefits to both methods, and together the two methods synergize their
-benefits in a unique way. The topology of convergence in distribution can be
-replicated with both the Wasserstein distance and the RKHS inner product, but
-the Wasserstein distance depends on far fewer parameters (such as the choice of
-the eponymous kernel in RKHS approaches), and its value is directly
-interpretable in terms of the shapes of the distributions it compares.
+We presented a general approach for predictive state analysis---Cantor fractal
+embedding sequences and Wasserstein distance comparison of predictions. We
+offered two approaches to visualizing the results of this method---one a direct
+application of multidimensional scaling and the other being a clustered Cantor
+diagram built from combining hierarchical clustering with the introduced Cantor
+embedding.
 
-Similarly, there are many ways to metrize the product topology on sequences, but
-the Cantor embedding offers a direct way to connect the product topology with a
-visualizable geometry, and by embedding in a single dimension enables efficient
-computation of the Wasserstein metric. The benefits of the Cantor and
-Wasserstein approaches adds interpretability to the resulting predictive state
-geometry along two distinct axes, most clearly seen in the clustered Cantor
-diagrams of Fig. \ref{fig:hclust}.
+Compared to using reproducing kernel Hilbert spaces---a dominant approach to
+predictive states at present
+\cite{Song09a,Song10a,Boot13a,Brod20a,Loom21a}---our combining the Cantor set
+with the Wasserstein distance may appear idiosyncratic. However, as the results
+demonstrated, there are strong benefits to both and together the two methods
+synergize their benefits in a unique way. The topology of convergence in
+distribution can be replicated with both the Wasserstein distance and the RKHS
+inner product. However, the Wasserstein distance depends on far fewer
+parameters---such as, the choice of the eponymous kernel in RKHS approaches.
+Moreover, its value is directly interpretable in terms of the shapes of the
+distributions it compares.
+
+Similarly, there are many ways to metrize the product topology on sequences,
+but the Cantor embedding offers a direct way to connect the product topology
+with a visualizable geometry. And, embedding in a single dimension enables
+efficient computation of the Wasserstein metric. The benefits of the Cantor and
+Wasserstein approaches adds interpretability to the resulting predictive-state
+geometry along two distinct axes, most clearly seen in Fig. \ref{fig:hclust}'s
+clustered Cantor diagrams.
 
@@
 hospitality during visits. This material is based upon work supported by, or
-  in part by, Grant Nos. FQXi-RFP-IPW-1902 and FQXi-RFP-1809 from the
-  Foundational Questions Institute and Fetzer Franklin Fund (a donor-advised
-  fund of Silicon Valley Community Foundation) and grants W911NF-18-1-0028 and
+in part by, Templeton World Charity Foundation Diverse Intelligences grants TWCF0440 and TWCF0570 and Foundational Questions Institute and Fetzer Franklin Fund grant number FQXI-RFP-CPW-2007 and grants W911NF-18-1-0028 and
 W911NF-21-1-0048 from the U.S. Army Research Laboratory and the U.S. Army
 Research Office.
+
 \end{acks}
@@
 \section{Reproducibility statement}
-For the purposes of reproducibility, we have provided a GitHub repository which
-contains the code necessary to generate this paper and its figures, including a
-notebook for generating the data we used for our examples and building both
+For the purposes of reproducibility, we provide a GitHub repository containing
+the code necessary to generate this manuscript and its figures, including a
+notebook for generating the data used for the examples and for building both
 static and interactive figures for further exploration.

### 2022-02-07 paper/manuscript.tex
final paper minus anonymized github link

@@
 length-$L$ sample $x_1\dots x_L$, and $n,k>0$, take a sliding window of pasts
-and futures, $(x_{t-n+1}\dots x_t, x_{t+1}\dots x_k)$ for $t=n,\dots,L-k$.  For
+and futures, $(x_{t-k+1}\dots x_t, x_{t+1}\dots x_n)$ for $t=k,\dots,L-n$.  For
 each past-future pair, compute the truncated Cantor embeddings on the
 \emph{reversed} past and (unreversed) future: $\left(C(x_t \dots
-x_{t-n+1}),C(x_{t+1}\dots x_k)\right)$. The resulting pairs of real numbers can
+x_{t-k+1}),C(x_{t+1}\dots x_n)\right)$. The resulting pairs of real numbers can
 be plotted as $(x,y)$-values on a scatter plot. The fractal that emerges
@@
 
-  \underline{function \CantorWasserstein} ($n,k,x_1\dots x_L$)\;
-  \Input{Integers $n,k$ of past and future lengths} 
+  \underline{function \CantorWasserstein} ($k,n,x_1\dots x_L$)\;
+  \Input{Integers $k,n$ of past and future lengths} 
   \Input{Length-$L$ sequence $x_1\dots x_L$ of observations} 
@@
   \For{$t\leftarrow n$ \KwTo $L-k$} { 
-    $\overleftarrow{\mathbf{x}} \leftarrow \left[x_{t-n+1},\dots, x_{t}\right]$\;
-    $\overrightarrow{\mathbf{x}} \leftarrow \left[x_{t+1},\dots, x_{t+k}\right]$\;
+    $\overleftarrow{\mathbf{x}} \leftarrow \left[x_{t-k+1},\dots, x_{t}\right]$\;
+    $\overrightarrow{\mathbf{x}} \leftarrow \left[x_{t+1},\dots, x_{t+n}\right]$\;
       \vspace{-1.5em}
       \begin{flalign*}
-        p &\leftarrow \sum_{\ell=1}^L
+        p &\leftarrow \sum_{\ell=1}^n
         \frac{2J(\overrightarrow{x}_\ell)}{(2|\mathcal{X}|-1)^\ell}\ ; &&
@@
   $\Wass \leftarrow \Matrix(K,K)$\;
-  \For{$k\leftarrow 1$ \KwTo $K$} { 
-    \For{$\ell\leftarrow 1$ \KwTo $K$} {  
-      $\Wass_{k \ell}
+  \For{$i\leftarrow 1$ \KwTo $K$} { 
+    \For{$j\leftarrow 1$ \KwTo $K$} {  
+      $\Wass_{ij}
         \leftarrow
-        \Wasserstein(\Cantors_k,\Cantors_\ell)$\;
-        $\Wass_{\ell k} \leftarrow \Wass_{k \ell}$\; } } 
+        \Wasserstein(\Cantors_i,\Cantors_j)$\;
+        $\Wass_{j i} \leftarrow \Wass_{i j}$\; } } 
   \KwResult{$\UnqPasts$,$\Cantors$,$\Wass$}
@@
 original logistic map. Said differently, there is a correspondence between each
-past $x_{-n+1}\dots x_{0}$ and a subset $V_{x_{-n+1}\dots x_{0}}$, such that
-$V_{x_1\dots x_n}$ is the set of all points $y$ for which $x(f^{-t}(y)) =
+past $x_{-k+1}\dots x_{0}$ and a subset $V_{x_{-k+1}\dots x_{0}}$, such that
+$V_{x_{-k+1}\dots x_{0}}$ is the set of all points $y$ for which $x(f^{-t}(y)) =
 x_{-t}$ for $0\leq t < n$. (Here, $f(y) = ry(1-y)$ and $x(y)$ is the encoding
-$y\mapsto \{0,1\}$.) As it happens, pasts $x_{-n+1}\dots x_{0}$ whose predictive
+$y\mapsto \{0,1\}$.) As it happens, pasts $x_{-k+1}\dots x_{0}$ whose predictive
 states are close under the Wasserstein distance are also pasts for which the
-sets $f(V_{x_{-n+1}\dots x_{0}})$ are close. That is, they correspond to
+sets $f(V_{x_{-k+1}\dots x_{0}})$ are close. That is, they correspond to
 predictively similar ranges of the logistic map variable.
@@
 
-Similarly, there are many ways to metrize the product topology on sequences,
-but the Cantor embedding offers a direct way to connect the product topology
-with a visualizable geometry. And, embedding in a single dimension enables
-efficient computation of the Wasserstein metric. The benefits of the Cantor and
+Similarly, there are many ways to metrize the product topology on sequences, but
+the Cantor embedding offers a direct way to connect the product topology with a
+visualizable geometry. And, embedding in a single dimension enables efficient
+computation of the Wasserstein metric. The benefits of the Cantor and
 Wasserstein approaches adds interpretability to the resulting predictive-state
 geometry along two distinct axes, most clearly seen in Fig. \ref{fig:hclust}'s
-clustered Cantor diagrams.
+clustered Cantor diagrams. We hope that the success of this approach in
+providing clear insights will complement existing thrusts in the direction of
+abstract embeddings and mathematical formalism by motivating further development
+on interpretable approaches to predictive state analysis.
 
@@
 \section{Reproducibility statement}
-For the purposes of reproducibility, we provide a GitHub repository containing
+For the purposes of reproducibility, we provide a 
+GitHub repository \cite{repo} containing
 the code necessary to generate this manuscript and its figures, including a

### 2022-06-01 paper/manuscript.tex
arxiv version

@@
 %% The first command in your LaTeX source must be the \documentclass command.
-\documentclass[sigconf, anonymous, review]{acmart}
+\documentclass[draft,aps,pre,twocolumn,groupaddress,showkeys,nofootinbib,preprintnumbers,floatfix]{revtex4-2}
 \usepackage{probs}
+\usepackage{dynlearn}
 \usepackage[linesnumbered,ruled]{algorithm2e}
 \mathchardef\mhyphen="2D
-%%
-%% \BibTeX command to typeset BibTeX logo in the docs
-\AtBeginDocument{%
-  \providecommand\BibTeX{{%
-    \normalfont B\kern-0.5em{\scshape i\kern-0.25em b}\kern-0.8em\TeX}}}
-
-%% Rights management information.  This information is sent to you % when you
-%complete the rights form.  These commands have SAMPLE % values in them; it is
-%your responsibility as an author to replace % the commands and values with
-%those provided to you when you % complete the rights form.
-\setcopyright{acmcopyright}
-\copyrightyear{2022}
-\acmYear{2022}
-%\acmDOI{10.1145/1122445.1122456}
-\acmConference[KDD'22]{the 28th ACM SIGKDD Conference on Knowledge Discovery and
-Data Mining}{August 2022}{Washington, DC, USA}
 
@@
 
-%%
-%% The "title" command has an optional parameter, % allowing the author to
-%define a "short title" to be used in page headers.
-\title{Predictive State Geometry via Cantor Embeddings and Wasserstein Distance}
-%%
-%% The "author" command and its associated commands are used to define % the
-%authors and their affiliations. % Of note is the shared affiliation of the
-%first two authors, and the % "authornote" and "authornotemark" commands % used
-%to denote shared contribution to the research.
+\def\ourTitle{%
+%  Analytical and statistical properties
+%  of predictive states
+Exploring Predictive States via Cantor Embeddings and Wasserstein Distance
+}
+
+\def\ourAbstract{%
+Predictive states for stochastic processes are a nonparametric and interpretable
+construct with relevance across a multitude of modeling paradigms. Recent
+progress on the self-supervised reconstruction of predictive states from
+time-series data focused on the use of reproducing kernel Hilbert spaces. Here,
+we examine how Wasserstein distances may be used to detect predictive
+equivalences in symbolic data. We compute Wasserstein distances between
+distributions over sequences (``predictions''), using a finite-dimensional
+embedding of sequences based on the Cantor for the underlying geometry. We show
+that exploratory data analysis using the resulting geometry via hierarchical
+clustering and dimension reduction provides insight into the temporal structure
+of processes ranging from the relatively simple (\emph{e.g.}, finite-state
+hidden Markov models) to the very complex (\emph{e.g.}, infinite-state indexed
+grammars).
+}
+
+\def\ourKeywords{%
+time series, predictive states, Wasserstein distance, hierarchical clustering
+}
+
+\hypersetup{
+  pdfauthor={Samuel Loomis},
+  pdftitle={\ourTitle},
+  pdfsubject={\ourAbstract},
+  pdfkeywords={\ourKeywords},
+  pdfproducer={},
+  pdfcreator={}
+}
+
 \author{Samuel P. Loomis}
 \email{sloomis@ucdavis.edu}
-\orcid{0000-0003-1587-9874}
 
@@
 \email{chaos@ucdavis.edu}
-\orcid{0000-0003-4466-5410}
-\affiliation{%
-  \institution{Complexity Sciences Center and Physics Department, \\University of California at Davis}
-  \streetaddress{One Shields Ave.}
-  \city{Davis}
-  \state{CA}
-  \country{USA}
-  \postcode{95616}
-}
+\affiliation{Complexity Sciences Center and Department of Physics and Astronomy,
+University of California at Davis, One Shields Avenue, Davis, CA 95616}
 
-%%
-%% By default, the full list of authors will be used in the page % headers.
-%Often, this list is too long, and will overlap % other information printed in
-%the page headers. This command allows % the author to define a more concise
-%list % of authors' names for this purpose.
-\renewcommand{\shortauthors}{Loomis and Crutchfield}
+\date{\today}
+\bibliographystyle{unsrt}
 
-%%
-%% The abstract is a short summary of the work to be presented in the % article.
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\title{\ourTitle}
+
 \begin{abstract}
-  Predictive states for stochastic processes are a nonparametric and
-  interpretable construct with relevance across a multitude of modeling
-  paradigms. Recent progress on the self-supervised reconstruction of predictive
-  states from time-series data focused on the use of reproducing kernel
-  Hilbert spaces. Here, we examine how Wasserstein distances may be used to
-  detect predictive equivalences in symbolic data. We construct a
-  finite-dimensional embedding of sequences based on the Cantor set and use
-  distances in this embedding to compute Wasserstein distances between
-  distributions over sequences (``predictions''). We show that analyzing the
-  resulting geometry via hierarchical clustering and dimension reduction
-  provides insight into the temporal structure of processes ranging from the
-  relatively simple (\emph{e.g.}, finite-state hidden Markov models) to the
-  very complex (\emph{e.g.}, infinite-state indexed grammars).
+\ourAbstract
 \end{abstract}
 
-%%
-%% The code below is generated by the tool at http://dl.acm.org/ccs.cfm. %
-%Please copy and paste the code instead of the example below.
-%%
-\begin{CCSXML}
-  <ccs2012>
-  <concept>
-  <concept_id>10002950.10003648.10003649</concept_id>
-  <concept_desc>Mathematics of computing~Probabilistic representations</concept_desc>
-  <concept_significance>500</concept_significance>
-  </concept>
-  <concept>
-  <concept_id>10002950.10003648.10003649.10003657</concept_id>
-  <concept_desc>Mathematics of computing~Nonparametric representations</concept_desc>
-  <concept_significance>500</concept_significance>
-  </concept>
-  <concept>
-  <concept_id>10002950.10003648.10003700.10003701</concept_id>
-  <concept_desc>Mathematics of computing~Markov processes</concept_desc>
-  <concept_significance>500</concept_significance>
-  </concept>
-  <concept>
-  <concept_id>10002950.10003648.10003700</concept_id>
-  <concept_desc>Mathematics of computing~Stochastic processes</concept_desc>
-  <concept_significance>500</concept_significance>
-  </concept>
-  <concept>
-  <concept_id>10002950.10003648.10003688.10003693</concept_id>
-  <concept_desc>Mathematics of computing~Time series analysis</concept_desc>
-  <concept_significance>500</concept_significance>
-  </concept>
-  <concept>
-  <concept_id>10002950.10003741.10003742</concept_id>
-  <concept_desc>Mathematics of computing~Topology</concept_desc>
-  <concept_significance>300</concept_significance>
-  </concept>
-  <concept>
-  <concept_id>10002950.10003712</concept_id>
-  <concept_desc>Mathematics of computing~Information theory</concept_desc>
-  <concept_significance>300</concept_significance>
-  </concept>
-  <concept>
-  <concept_id>10003752.10003766</concept_id>
-  <concept_desc>Theory of computation~Formal languages and automata theory</concept_desc>
-  <concept_significance>300</concept_significance>
-  </concept>
-  <concept>
-  <concept_id>10003752.10003766.10003771</concept_id>
-  <concept_desc>Theory of computation~Grammars and context-free languages</concept_desc>
-  <concept_significance>300</concept_significance>
-  </concept>
-  </ccs2012>
-\end{CCSXML}
+\keywords{\ourKeywords}
 
-\ccsdesc[500]{Mathematics of computing~Probabilistic representations}
-\ccsdesc[500]{Mathematics of computing~Nonparametric representations}
-\ccsdesc[500]{Mathematics of computing~Stochastic processes}
-\ccsdesc[500]{Mathematics of computing~Time series analysis}
-\ccsdesc[300]{Mathematics of computing~Markov processes}
-\ccsdesc[300]{Mathematics of computing~Topology}
-\ccsdesc[300]{Mathematics of computing~Information theory}
-\ccsdesc[300]{Theory of computation~Formal languages and automata theory}
-\ccsdesc[300]{Theory of computation~Grammars and context-free languages}
+% \pacs{
+% 05.45.-a  %  Nonlinear dynamics and nonlinear dynamical systems
+% 89.75.Kd  %  Complex Systems: Patterns
+% 89.70.+c  %  Information science
+% 05.45.Tp  %  Time series analysis
+% 02.50.Ey  %  Stochastic processes
+% 02.50.-r  %  Probability theory, stochastic processes, and statistics
+% 02.50.Ga  %  Markov processes
+% 05.20.-y  %  Classical statistical mechanics
+% }
 
-%%
-%% Keywords. The author(s) should pick words that accurately describe % the work
-%being presented. Separate the keywords with commas.
-\keywords{time series, predictive states, Cantor set, Wasserstein distance, hierarchical clustering}
+\preprint{\arxiv{2108.XXXXX}}
 
-%% A "teaser" image appears between the author and affiliation % information and
-%the body of the document, and typically spans the % page.
+\date{\today}
+\maketitle
 
-%%
-%% This command processes the author and affiliation and title % information and
-%builds the first part of the formatted document.
-\maketitle
+\tableofcontents
+% Clears up font spacing/overlap in ToC
+% with sufficient repetitions of formating (e.g., pdflatex).
+% Other front matter packages can alter formating constants.
+\makeatletter
+\let\toc@pre\relax
+\let\toc@post\relax
+\makeatother
+
+\setstretch{1.1}
+
+\listoffixmes
 
@@
 theory of predictive-state inference has been correspondingly developed
-\cite{Jaeg00a,Shal02a,Jame04a,Stil07b,Stre13a,Thon15a,Marz19a,Brod20a}. Despite
-this, universal conditions for predictive-state convergence were not given until
-recently. It is now known that if the dataset $x_1 \dots x_L$ is drawn from any
+\cite{Uppe97a,Jaeg00a,Shal02a,Jame04a,Stil07b,Stre13a,Thon15a,Marz19a,Brod20a}.
+If the dataset $x_1 \dots x_L$ is drawn from any
 stationary process and if $L$ is sufficiently large, then Eq.
 \eqref{eq:convergence} converges for any $n$ and a probability-$1$ subset of
-pasts $\overleftarrow{x}$ \cite{Loom21a}. In the language of measures,
+pasts $\overleftarrow{x}$ \cite{Uppe97a}. Recently, this result has been further
+clarified: in the language of measures,
 $P_{\overleftarrow{x}}$ converges \emph{in distribution}, with respect to the
-\emph{product topology} of the space of sequences $\mathcal{X}^\mathbb{N}$.
+\emph{product topology} of the space of sequences $\mathcal{X}^\mathbb{N}$
+\cite{Loom21a}.
 
@@
 understand the dynamics of how predictions evolve under a stream of new
-observations. The following focuses on the first, as it is a crucial building
-block to achieving the other two.
+observations.
 
-Recent attempts to reconstruct the geometry of predictive states embedded them
+The following focuses on the first, as it is a crucial building
+block to achieving the other two. Recent attempts to reconstruct the geometry of predictive states embedded them
 in reproducing kernel Hilbert spaces
@@
 they respect the product topology of sequences in the same way that predictive
-states themselves do.
+states themselves do \cite{Loom21a}. In this paper, we will seek to achieve the same results,
+using a procedure which can be easily visualized and interpreted, making it
+suitable as a method for exploratory data analysis \cite{Tuke62a,Tuke77a}.
 
@@
 	even size bounded by $0$s. The even process has infinite Markov order but
-	can be generated by a two-state hidden Markov chain \cite{Crut01a}.
+	can be generated by a two-state hidden Markov chain \cite{Crut01a}. A typical
+	example might look like $01100111101100011$.
 \item The $a^n b^n$ \emph{process} can be generated by choosing a random
@@
 	by any finite hidden Markov chain, though it is a simple example of a
-	probabilistic context-free language \cite{Hopc06a}.
+	probabilistic context-free language \cite{Hopc06a}. A typical example might
+	look like $\mathtt{abaaabbbabaabb}$.
 \item The $x+f(x)$ \emph{process} is a probabilistic context-free language
@@
 \end{align*}
+A typical example might
+	look like $\mathtt{x+x;f(x+f(x));f(f(x));}$.
 \item The $a^n b^n c^n$ \emph{process} is a probabilistic indexed language
@@
 	after writing the blocks of $\mathtt{a}$'s and $\mathtt{b}$'s, we also
-	write a block of $\mathtt{c}$'s of length $n$.
+	write a block of $\mathtt{c}$'s of length $n$. A typical example might
+	look like $\mathtt{abcaaabbbcccabcaabbcc}$.
 \item The \emph{Morse-Thue process} is generated by sampling from the time
@@
 	consecutively. The resulting process is an indexed-context free language
-	\cite{Crut92c}.
+	\cite{Crut92c}. A typical example might look like
+	$11011101010111011101110101011101$ (this is the fifth generation of the
+	replacement rule starting from $0$).
 \end{enumerate}
 
-\begin{figure*}[h]
+\begin{figure*}[ht]
 \centering
@@
 \label{fig:cantor}
-\Description{Four fractals are displayed, respectively labeled as the even,
-	$\mathtt{a}^n\mathtt{b}^n$, $\mathtt{a}^n\mathtt{b}^n\mathtt{c}^n$, and
-	$\mathtt{x+f(x)}$ processes.
-	}
 \end{figure*}
@@
 as to not outweigh the computational benefits of working in one dimension.
-
 \begin{algorithm}[t]
@@
 \label{fig:hclust}
-\Description{Four graphs are displayed. Each vertical axis has a list of
-	symbolic sequences organized by a hierarchical clustering tree. The
-	horizontal axis is a list of individual symbols. The content of each graph
-	is rows of fractal-distributed lines.
-	}
 \end{figure*}
@@
 \label{fig:mds}
-\Description{Two scatter plots with colored clusters of points are displayed,
-	with labels next to each cluster describing symbolic patterns.
-	}
 \end{figure*}
@@
 
-\begin{figure*}[h]
+\begin{figure*}[ht]
 \centering
@@
 \label{fig:mds-feig}
-\Description{A rectangular scatter plot of color-coded points is displayed on
-	the left, and on the right a square scatterplot of parabolically arranged
-	points with a diagonal line cutting through them. In the right plot the
-	points are colored with the same scheme as the left, and similarly colored
-	arrows point from clusters to the diagonal line or from the diagonal line
-	to the next cluster.
-	}
 \end{figure*}
@@
 %section in the article metadata, and the % consistent spelling of the heading.
-\begin{acks}
+\section*{Acknowledgements}
 We thank Nicolas Brodu, Adam Rupe, Alex Jurgens, David Gier, and Mikhael
@@
 
-\end{acks}
-
-\section{Reproducibility statement}
+\section*{Reproducibility statement}
 For the purposes of reproducibility, we provide a 
@@
 %bibliography file.
-\bibliographystyle{ACM-Reference-Format}
 \bibliography{ref}

### 2022-06-06 paper/manuscript.tex
JPC: Started editing.

@@
-%%
-%% This is file `sample-lualatex.tex', % generated with the docstrip utility.
-%%
-%% The original source files were:
-%%
-%% samples.dtx  (with options: `sigconf')
-%% 
-%% IMPORTANT NOTICE:
-%% 
-%% For the copyright see the source file.
-%% 
-%% Any modified versions of this file must be renamed % with new filenames
-%distinct from sample-lualatex.tex.
-%% 
-%% For distribution of the original source see the terms % for copying and
-%modification in the file samples.dtx.
-%% 
-%% This generated file may be distributed as long as the % original source
-%files, as listed above, are part of the % same distribution. (The sources need
-%not necessarily be % in the same archive or directory.)
-%%
-%%
+% Exploring Predictive States via Cantor Embeddings and Wasserstein Distance
+%
+% spl: 6/2/22
+% jpc: 6/5/22
+
 %% Commands for TeXCount TC:macro \cite [option:text,text] TC:macro \citep
@@
 %%
-%%
-%% The first command in your LaTeX source must be the \documentclass command.
+
 \documentclass[draft,aps,pre,twocolumn,groupaddress,showkeys,nofootinbib,preprintnumbers,floatfix]{revtex4-2}
@@
 
-\preprint{\arxiv{2108.XXXXX}}
+\preprint{\arxiv{2206.XXXXX}}
 
@@
 
-\tableofcontents
+% \tableofcontents
 % Clears up font spacing/overlap in ToC
@@
 stationary (time-translation invariant) and ergodic (explores
-all possible behaviors) \cite{Kall01a,Shal07a}. We may wish to forecast from the
+all possible behaviors) \cite{Fell70a,Kall01a}. We may wish to forecast from the
 observed information the behavior of the next $n$ observations. If we suspect
@@
 
-Predictive states are employed for inference and modeling in dynamical
-systems \cite{Crut92c,Emen16a,Salo19a}, renewal processes and spike-trains
-\cite{Marz14e,Marz15a}, condensed matter physics \cite{Varn12a,Varn14a}, and
-spatiotemporal systems \cite{Rupe19a,Rupe19b,Rupe20a}. A deep mathematical
+Predictive states are employed for inference and modeling in dynamical systems
+\cite{Crut92c}, renewal processes and neural spike-trains
+\cite{Marz14e,Marz15a}, condensed matter physics \cite{Varn14a}, and
+spatiotemporal systems \cite{Rupe19b}. A deep mathematical
 theory of predictive-state inference has been correspondingly developed
-\cite{Uppe97a,Jaeg00a,Shal02a,Jame04a,Stil07b,Stre13a,Thon15a,Marz19a,Brod20a}.
-If the dataset $x_1 \dots x_L$ is drawn from any
-stationary process and if $L$ is sufficiently large, then Eq.
-\eqref{eq:convergence} converges for any $n$ and a probability-$1$ subset of
-pasts $\overleftarrow{x}$ \cite{Uppe97a}. Recently, this result has been further
-clarified: in the language of measures,
+\cite{Uppe97a,Jaeg00a,Shal02a,Jame04a,Stil07b,Stre13a,Thon15a,Brod20a}.
+If the dataset $x_1 \dots x_L$ is drawn from any stationary process and if $L$
+is sufficiently large, then Eq. \eqref{eq:convergence} converges for any $n$
+and a probability-$1$ subset of pasts $\overleftarrow{x}$ \cite{Uppe97a}.
+Recently, this result has been further clarified: in the language of measures,
 $P_{\overleftarrow{x}}$ converges \emph{in distribution}, with respect to the
@@
 
-The following focuses on the first, as it is a crucial building
-block to achieving the other two. Recent attempts to reconstruct the geometry of predictive states embedded them
-in reproducing kernel Hilbert spaces
+The following focuses on the first, as it is a crucial building block to
+achieving the other two. Recent attempts to reconstruct the geometry of
+predictive states embedded them in reproducing kernel Hilbert spaces
 \cite{Song09a,Song10a,Boot13a,Brod20a,Loom21a}. This was achieved to great
@@
 they respect the product topology of sequences in the same way that predictive
-states themselves do \cite{Loom21a}. In this paper, we will seek to achieve the same results,
-using a procedure which can be easily visualized and interpreted, making it
-suitable as a method for exploratory data analysis \cite{Tuke62a,Tuke77a}.
+states themselves do \cite{Loom21a}. Here, we achieve the same results using a
+procedure that can be easily visualized and interpreted. This makes it suitable
+as a method for exploratory data analysis \cite{Tuke62a,Tuke77a}, while moving
+further toward interpretable machine learning of structured processes.
 
@@
 
-We refer back frequently to the following example processes:
+We refer back frequently to the following example processes of increasing
+computational complexity:
 \begin{enumerate}
+      \setlength{\topsep}{-2pt}
+      \setlength{\itemsep}{-2pt}
+      \setlength{\parsep}{-2pt}
 \item The \emph{even process} can be generated by repeatedly tossing a coin
@@
 	even size bounded by $0$s. The even process has infinite Markov order but
-	can be generated by a two-state hidden Markov chain \cite{Crut01a}. A typical
-	example might look like $01100111101100011$.
-\item The $a^n b^n$ \emph{process} can be generated by choosing a random
-	integer $n\geq 1$ (we suppose via a Poisson process) and writing $n$
-	$\mathtt{a}$'s followed by an equal number of $\mathtt{b}$'s, and then
-	repeating this procedure indefinitely. This results in sequences where any
-	contiguous block of $\mathtt{a}$s is followed by a block of $\mathtt{b}$s
-	of equal size. The $\mathtt{a}^n \mathtt{b}^n$ process cannot be generated
-	by any finite hidden Markov chain, though it is a simple example of a
-	probabilistic context-free language \cite{Hopc06a}. A typical example might
-	look like $\mathtt{abaaabbbabaabb}$.
+	can be generated by a two-state hidden Markov chain \cite{Crut01a}. A
+	typical example might look like $01100111101100011$.
+\item The $\mathtt{a}^n \mathtt{b}^n$ \emph{process} can be generated by
+	choosing a random integer $n\geq 1$ (we suppose via a Poisson process) and
+	writing $n$ $\mathtt{a}$s followed by an equal number of $\mathtt{b}$s,
+	and then repeating this procedure indefinitely. This results in sequences
+	where any contiguous block of $\mathtt{a}$s is followed by a block of
+	$\mathtt{b}$s of equal size. The $\mathtt{a}^n \mathtt{b}^n$ process cannot
+	be generated by any finite hidden Markov chain, though it is a simple
+	example of a probabilistic context-free language \cite{Hopc06a}. A typical
+	example might look like $\mathtt{abaaabbbabaabb}$.
 \item The $x+f(x)$ \emph{process} is a probabilistic context-free language
@@
 \end{align*}
-A typical example might
-	look like $\mathtt{x+x;f(x+f(x));f(f(x));}$.
-\item The $a^n b^n c^n$ \emph{process} is a probabilistic indexed language
-	\cite{Hopc06a} that is analogous to $\mathtt{a}^n \mathtt{b}^n$ except
-	after writing the blocks of $\mathtt{a}$'s and $\mathtt{b}$'s, we also
-	write a block of $\mathtt{c}$'s of length $n$. A typical example might
-	look like $\mathtt{abcaaabbbcccabcaabbcc}$.
+	A typical example might look like $\mathtt{x+x;f(x+f(x));f(f(x));}$.
+\item The $\mathtt{a}^n \mathtt{b}^n \mathtt{c}^n$ \emph{process} is a
+	probabilistic indexed language \cite{Hopc06a} that is analogous to
+	$\mathtt{a}^n \mathtt{b}^n$ except after writing the blocks of
+	$\mathtt{a}$'s and $\mathtt{b}$'s, we also write a block of $\mathtt{c}$'s
+	of length $n$. A typical example might look like
+	$\mathtt{abcaaabbbcccabcaabbcc}$.
 \item The \emph{Morse-Thue process} is generated by sampling from the time
@@
 	\cite{Crut92c}. A typical example might look like
-	$11011101010111011101110101011101$ (this is the fifth generation of the
-	replacement rule starting from $0$).
+	$11011101010111011101110101011101$---the fifth generation of the
+	replacement rule starting from $0$.
 \end{enumerate}
@@
 \section{Cantor-embedding sequences}
+
+STOPPED
 
@@
   \KwResult{$\UnqPasts$,$\Cantors$,$\Wass$}
-  \caption{Steps for converting a sequence of categorical time-series data
-        into a labeled collection of empirical distributions of Cantor-embedded futures,
-        and a matrix of Wasserstein distances from said distributions.}
+\caption{Convert a sequence of categorical time-series data into a labeled
+	collection of empirical distributions of Cantor-embedded futures and a
+	matrix of Wasserstein distances between said distributions.
+	}
 \label{alg:cantorwass}

### 2022-06-08 paper/manuscript.tex
JPC: Editing.
JPC: Editing almost done.

@@
 \centering
-\includegraphics[width=0.4\linewidth]{../plots/even_CANTOR.png}
-\includegraphics[width=0.4\linewidth]{../plots/anbn_CANTOR.png}
-\includegraphics[width=0.4\linewidth]{../plots/anbncn_CANTOR.png}
-\includegraphics[width=0.4\linewidth]{../plots/x+f(x)_CANTOR.png}
+\includegraphics[width=0.46\linewidth]{../plots/even_CANTOR.png}
+\includegraphics[width=0.45\linewidth]{../plots/anbn_CANTOR.png}
+\includegraphics[width=0.45\linewidth]{../plots/anbncn_CANTOR.png}
+\includegraphics[width=0.45\linewidth]{../plots/x+f(x)_CANTOR.png}
 \caption{Cantor plots for the even, $\mathtt{a}^n\mathtt{b}^n$,
@@
 	Each point $(x,y)$ corresponds to a pair of sequences corresponding to the
-	past and future, respectively. The symbol on the $x$ ($y$) axis indicates
-	that all points above (to the right of) that symbol have a past (future)
-	whose most recent observation is that symbol. Though not marked, further
-	proportional subdivisions of each segment of the axes indicate the value of
-	the second, third, \emph{etc.} symbols. For instance, one can read from the
-	$\mathtt{x+f(x)}$ fractal that any past ending in $\mathtt{f}$ must be
-	paired with a future beginning in $\mathtt{(f}$ or $\mathtt{(x}$.
+	past $x$ and future $y$, respectively. The symbol on the $x$ ($y$) axis
+	indicates that all points above (to the right of) that symbol have a past
+	(future) whose most recent observation is that symbol. Though not marked,
+	further proportional subdivisions of each segment of the axes indicate the
+	value of the second, third, and so on symbols. For instance, one can read
+	from the $\mathtt{x+f(x)}$ fractal that any past ending in $\mathtt{f}$
+	must be paired with a future beginning in $\mathtt{(f}$ or $\mathtt{(x}$.
 	}
@@
 \section{Cantor-embedding sequences}
-
-STOPPED
 
@@
 It is the minimal cost to ``shift'' the probability mass from one distribution
-to match shape of the other.
+to match the other's shape.
 
@@
 \centering
-\includegraphics[width=0.45\linewidth]{../plots/even_HCLUST.png}
-\includegraphics[width=0.45\linewidth]{../plots/anbn_HCLUST.png}
-\includegraphics[width=0.45\linewidth]{../plots/anbncn_HCLUST.png}
-\includegraphics[width=0.45\linewidth]{../plots/x+f(x)_HCLUST.png}
+\includegraphics[width=0.49\linewidth]{../plots/even_HCLUST.png}
+\includegraphics[width=0.49\linewidth]{../plots/anbn_HCLUST.png}
+\includegraphics[width=0.49\linewidth]{../plots/anbncn_HCLUST.png}
+\includegraphics[width=0.49\linewidth]{../plots/x+f(x)_HCLUST.png}
 \caption{(Upper left to lower right) Clustered Cantor diagrams of the even,
@@
 The resulting clustered Cantor plots offer a highly interpretable visualization
-of the relationship between pasts and futures, and of the predictive states'
+of the relationship between pasts and futures and of the predictive states'
 geometry. Each plot, in a certain sense, sorts the columns in the Cantor
@@
 and $\mathtt{x+f(x)}$ processes as well. We draw attention to the manner in
-which the presence of a semicolon in pasts from $\mathtt{x+f(x)}$ affects the
-comparison of predictions.
+which the presence of a semicolon in pasts from the $\mathtt{x+f(x)}$ process
+affects the comparison of predictions.
 
@@
 \centering
-\includegraphics[width=0.45\linewidth]{../plots/even_MDS.pdf}
-\includegraphics[width=0.45\linewidth]{../plots/anbn_MDS.pdf}
+\includegraphics[width=0.49\linewidth]{../plots/even_MDS.pdf}
+\includegraphics[width=0.48\linewidth]{../plots/anbn_MDS.pdf}
 \caption{Scatterplots of the first two MDS coordinates of the reconstructed
@@
 prediction clusters. These correspond to the predictive states that result
-from seeing an even-sized block of $1$s (or, equivalently, no $1$s), and that
-result from seeing an odd-sized block of $1$s.
+from seeing an even-sized block of $1$s (or, equivalently, no $1$s) and that
+result from seeing an odd-sized block of $1$s. The lone cluster in the middle
+corrresponds to a transient state induced by seeing all $1$s and then a $0$.
+The latter then synchronizes to right cluster.
 
@@
 Recall that the Morse-Thue process is a coarse-graining of the iterated
-logistic map $y_{t+1} = ry_t(1-y_t)$ at the critical chaos parameter $r_c
+logistic map $y_{t+1} = ry_t(1-y_t), ~y_0 = 0,$ at the critical chaos parameter $r_c
 \approx 3.56995$. The resulting stream of $0$s and $1$s is a well-known
@@
 \section{Concluding remarks}
+
+STOPPED
 

### 2022-06-08 paper/manuscript.tex
JPC: Done with edit pass. Now to submit.

@@
 result from seeing an odd-sized block of $1$s. The lone cluster in the middle
-corrresponds to a transient state induced by seeing all $1$s and then a $0$.
+corresponds to a transient state induced by seeing all $1$s and then a $0$.
 The latter then synchronizes to right cluster.
@@
 
-STOPPED
-
 We presented a general approach for predictive state analysis---Cantor fractal
@@
 %section in the article metadata, and the % consistent spelling of the heading.
-\section*{Acknowledgements}
+\vspace{-0.2in}
+\section*{Acknowledgments}
+\vspace{-0.2in}
 We thank Nicolas Brodu, Adam Rupe, Alex Jurgens, David Gier, and Mikhael
@@
 
+\vspace{-0.2in}
 \section*{Reproducibility statement}
+\vspace{-0.2in}
 For the purposes of reproducibility, we provide a 
@@
 %bibliography file.
+
 \bibliography{ref}

### 2022-06-11 paper/manuscript.tex
JPC: Added Lead Paragraph for Chaos.

@@
 \listoffixmes
+
+{\bf
+\section*{Lead:}
+Discovering the hidden structure of complex systems has been a widely-recognized
+goal of nonlinear science for decades, originally starting with extracting
+``geometry'' from a time series. The present results follow directly in this
+long line of inquiry, especially that focusing on the causal states or
+effective theories underlying time series generated by complex nonlinear
+systems. It provides workable statistical estimation methods and dynamical
+interpretations of hidden mechanisms. It establishes a new mathematical
+foundation for causal inference of complex systems, partly by making
+connections to modern machine learning. In this, its use of Cantor embeddings and Wasserstein distances complements recent work on reproducing-kernel Hilbert space representations.
+}
 

### 2022-09-19 paper/manuscript.tex
JPC: Updated funding acks.

@@
 hospitality during visits. This material is based upon work supported by, or
-in part by, Templeton World Charity Foundation Diverse Intelligences grants TWCF0440 and TWCF0570 and Foundational Questions Institute and Fetzer Franklin Fund grant number FQXI-RFP-CPW-2007 and grants W911NF-18-1-0028 and
-W911NF-21-1-0048 from the U.S. Army Research Laboratory and the U.S. Army
-Research Office.
+in part by, Templeton World Charity Foundation Diverse Intelligences grants
+TWCF0440 to the SETI Institute and TWCF0570 to UC Davis and by Foundational
+Questions Institute and Fetzer Franklin Fund grant number FQXI-RFP-CPW-2007 and
+grants W911NF-18-1-0028 and W911NF-21-1-0048 from the U.S. Army Research
+Laboratory and the U.S. Army Research Office.
 

### 2022-09-28 paper/manuscript.tex
paper edits

@@
 distributions over sequences (``predictions''), using a finite-dimensional
-embedding of sequences based on the Cantor for the underlying geometry. We show
+embedding of sequences based on the Cantor set for the underlying geometry. We show
 that exploratory data analysis using the resulting geometry via hierarchical
@@
 systems. It provides workable statistical estimation methods and dynamical
-interpretations of hidden mechanisms. It establishes a new mathematical
+interpretations of hidden mechanisms. It expands upon a new mathematical
 foundation for causal inference of complex systems, partly by making
@@
 achieving the other two. Recent attempts to reconstruct the geometry of
-predictive states embedded them in reproducing kernel Hilbert spaces
+predictive states embedded them in reproducing kernel Hilbert spaces (RKHS)
 \cite{Song09a,Song10a,Boot13a,Brod20a,Loom21a}. This was achieved to great
@@
       \setlength{\parsep}{-2pt}
+\item The \emph{fair coin process} is simply generated by flipping a coin repeatedly and
+writing down a $0$ for every tail and $1$ for every head. There is no memory in
+this process, and all pasts will map to the same predictive state. We will
+discuss it occasionally as a counterpoint to the more complex examples below.
+We also provide code and figures for this example in our GitHub repository
+\cite{repo}. 
 \item The \emph{even process} can be generated by repeatedly tossing a coin
@@
 	\mathtt{+}\ ,\ \mathtt{f}\ ,\ \mathtt{x}\}$ and nonterminals $\{A,B,C\}$,
-	and starts with a sequence of $A$s. Sequences are generated by applying the
+	and starts with a finite sequence of $A$s (such as $AAA$). Sequences are generated by applying the
 	production rules:
@@
 \end{align*}
-	A typical example might look like $\mathtt{x+x;f(x+f(x));f(f(x));}$.
+	The sequence generation process does not ``end'' until all non-terminals
+	(that is, $A$'s, $B$'s and $C$'s) have been replaced by terminal symbols.
+	A typical example might look like $\mathtt{x+x;f(x+f(x));f(f(x));}$ which
+	results from the substitutions
+	\begin{align*}
+		AAA &\mapsto B\ \mathtt{+}\ C\ \mathtt{;}
+		\ C\ \mathtt{;}\ C\ \mathtt{;}\\
+		&\mapsto C\mathtt{+x;}
+		\mathtt{f(}B\mathtt{);}\mathtt{f(}B\mathtt{);}\\
+		&\mapsto \mathtt{x+x;f(}B\ \mathtt{+}\ C\mathtt{);f(}C\mathtt{);}\\
+		&\mapsto \mathtt{x+x;f(}C\ \mathtt{+f(}C\mathtt{));f(f(}C\mathtt{));}\\
+		&\mapsto \mathtt{x+x;f(x+f(x));f(f(x));}
+	\end{align*}
 \item The $\mathtt{a}^n \mathtt{b}^n \mathtt{c}^n$ \emph{process} is a
@@
 	of length $n$. A typical example might look like
-	$\mathtt{abcaaabbbcccabcaabbcc}$.
+	$\mathtt{abcaaabbbcccabcaabbcc}$. Though it looks very similar to the
+	$\mathtt{a}^n \mathtt{b}^n$ process, it is in a formally distinct complexity
+	class, as $\mathtt{a}^n \mathtt{b}^n$ may be generated with a simple stack
+	automaton while $\mathtt{a}^n \mathtt{b}^n \mathtt{c}^n$ requires nested stacks.
 \item The \emph{Morse-Thue process} is generated by sampling from the time
@@
 \end{enumerate}
+
+Though our examples are abstract, they represent real structures which may arise
+in practice.
+
+Examples 3 through 5, in particular, involve long-range
+correlations in their sequences which arise from structures which, upon
+``opening,'' must be ``closed'' an arbitrary time later. Such structures are
+present in human language as well as in genetic sequences. In the latter case,
+this long-range correlation exists due to the folding of proteins which allows
+distant codons to interact with one another. Detecting and modeling these
+folding structures requires
+more complex models than hidden Markov models, such as the context-free grammars
+in examples 3 and 4 \cite{Saka94a}. 
+Example 6 represents the general problem of recovering knowledge of a dynamical
+system of which only coarse-grained data is available.
 
@@
 $x_1 x_2 \dots x_L$ and a following sequence $\overrightarrow{x}_L =
-(x_L,x_{L+1},\dots)$. That is, the space of sequences $\mathcal{X}^\mathbb{N}$
-can be factored into $\mathcal{X}^L\times \mathcal{X}^\mathbb{N}$ for any $L$.
+(x_{L+1},x_{L+2},\dots)$. That is, the space of sequences $\mathcal{X}^\mathbb{N}$
+can be factored into $\mathcal{X}^L\times \mathcal{X}^\mathbb{N}$ for any $L$;
+much like a fractal, then, $\mathcal{X}^\mathbb{N}$ is comprised of copies of itself.
 The fractal nature of sequence-space is encoded in the structure of its
-\emph{product topology}.
+\emph{product topology}, in which the closeness of two sequences is measured by
+the number of sites at which their values match. In other words, the
+neighborhoods of the product topology are generated by these smaller copies of
+$\mathcal{X}^\mathbb{N}$.
 
-We exploit this self-similarity in an interesting way by constructing a mapping
+We can visualize this self-similarity in an interesting way by constructing a mapping
 between sequence space and the celebrated \emph{Cantor set} (or one of its
@@
 \end{align*}
-For instance, suppose that $|\mathcal{X}|=2$ has two elements; then $C$ maps the
-sequence to a point the traditional Cantor set fractal. For a finite sequence of
-length $L$, truncate the sum at $k=L$.
+For instance, suppose that $|\mathcal{X}|=2$ has two elements; then the mapping
+$C$ looks like
+\begin{align*}
+	C(x_1,x_2,\dots) = \sum_{k=1}^\infty \frac{2J(x_k)}{3^k}
+	~.
+  \end{align*}
+The range of this mapping, taken over all binary sequences, matches the
+traditional Cantor set fractal, which is attained by removing the middle third
+from the unit interval, and repeating this process on all remaining intervals
+\emph{ad infinitum}. When $\mathcal{X}$ has more than two elements, the
+resulting range is a generalized form of the Cantor set, where instead of
+removing the middle (second) third, we remove the second and fourth quintiles ($|\mathcal{X}|=3$), or
+second, fourth and sixth heptiles ($|\mathcal{X}|=4$), and so on.
+For a finite sequence of
+length $L$, we truncate the sum at $k=L$, resulting in a finite-depth
+approximation to the Cantor set.
 
-Remarkably, the embedding $C$ has the property that for any continuous function
-$f$ on $[0,1]$, the function $F(\overrightarrow{x}) = f(C(\overrightarrow{x}))$
-is continuous on $\mathcal{X}^\mathbb{N}$. Further, if $F$ is continuous on
-$\mathcal{X}^\mathbb{N}$, then $f(y) = F(C^{-1}(y))$ is continuous on the image.
-Thus, the embedding $C$ respects the basic structure of the product topology
-\cite{Kurk03a}.
+Why use this somewhat convoluted construction for an embedding? The reason is
+that the embedding $C$ preserves the basic structure of the product topology on
+sequences \cite{Kurk03a}. Imagine, for instance, that we chose a more direct
+map, such as mapping a binary sequence to the corresponding binary number
+(\emph{e.g.} $010110\dots \mapsto 0.010110\dots$): then dissimilar sequences
+such as $0111\dots$ and $1000\dots$ would in fact map to the same real number.
+The padding added by removed intervals in the Cantor set means we can map distinct sequences to distinct real
+numbers, while keeping similar sequences close together.
 
@@
 structure that may or may not be desired. Associating each symbol $x$ with an
-integer $j_x$ endows an ordinal structure on the set $\mathcal{X}$. This
-ordinality is present in the macroscopic geometry of
-$C\left(\mathcal{X}^{\mathbb{N}}\right)$. Later, we examine higher-dimension
-embeddings that do not have this. However, they come at the cost of increased
+integer $J(x)$ endows an ordinal structure on the set $\mathcal{X}$. That is,
+the specific mapping from symbols to integers is arbitrary, and the way the
+symbols are ordered in the Cantor embedding---for instance, which symbol gets
+assigned to the middle segment instead of the end segments when
+$|\mathcal{X}|=3$---can affect the results of the visualization and downstream
+processing. Higher-dimension embeddings may be contrived which do not impose an
+ordering on the symbols. However, they come at the cost of increased
 computational complexity when determining the Wasserstein distance. So, for now,
-we assume that ordinal artifacts are either desired or sufficiently tolerable
-as to not outweigh the computational benefits of working in one dimension.
+we assume that ordinal artifacts are either desired or sufficiently tolerable as
+to not outweigh the computational benefits of working in one dimension.
 \begin{algorithm}[t]
@@
 column is just one of two types. This corresponds with the $2$ states of the
-hidden Markov model that generates the even process.
+hidden Markov model that generates the even process. By comparison, if we had
+plotted the fair coin's Cantor fractal, we would have found it to be perfectly
+symmetric, with every vertical column identical.
 
@@
 
-The intuitive distance metric between probability measures for capturing this
-underlying geometry is the Wasserstein metric \cite{Pane19a}. Given two measures
+This provides some intuition for how to effectively choose a distance metric for
+predictive states. Metrics such as the total variation distance or
+divergence-based distance metrics are too coarse; a small shift in the support
+of the distributions can result in dramatic changes in the distances between the
+distributions. Instead, we want a distance metric for distributions that
+quantifies the differences between the supports. The Wasserstein metric, also known as
+the earth-mover's distance, provides this feature \cite{Pane19a}. As the name ``earth-mover's''
+suggests, we may imagine the probability distributions as masses of dirt, and
+quantify the difference between two distribution as the cost of shifting the
+dirt around to transform one distribution to another. This way, if two
+distributions have similar but non-overlapping supports, the Wasserstein metric
+will be small.
+
+Formally, we can define the Wasserstein metric as follows. Given two measures
 $\mu$ and $\nu$ defined on a metric space $\mathcal{M}$ with metric $d$, the
@@
 
-$W(\mu,\nu)$ is the solution to a constrained linear optimization. As a
-function of distributions, $W(\mu,\nu)$ is continuous with respect to
-convergence in distribution. In fact, convergence under the Wasserstein
-distance is equivalent to convergence in distribution on compact spaces. This
-makes $W(\mu,\nu)$ ideal for measuring geometry between predictive states,
-since empirical estimates of these are known to converge in distribution.
-
-When $\mathcal{M} \subseteq \mathbb{R}$, there is in fact a closed-form solution
+$W(\mu,\nu)$ is the solution to a constrained linear optimization (since the
+objective function and constraints are linear functions of $\pi$). This can
+become computationally costly, scaling as $O(n^3 \log n)$ in the number of
+samples $n$ \cite{Pele09a}.
+However, when $\mathcal{M} \subseteq \mathbb{R}$, there is in fact a closed-form solution
 to the Wasserstein optimization problem \cite{Thas10a}. Let $F$ and $G$
@@
 this formula.
+
+Topologically, the Wasserstein distance replicates the topology of convergence
+in distribution, which means that algorithms which construct predictive states using the
+Wasserstein distance will converge.
 
@@
 scaling (MDS) decomposition \cite{Borg05a} for the even and
-$\mathtt{a}^n\mathtt{b}^n$ processes. Clusters are colored in the same manner
-as in Fig. \ref{fig:hclust} and labeled by the specific pattern that
+$\mathtt{a}^n\mathtt{b}^n$ processes. MDS extracts coordinates so that the
+Euclidean distance between points in MDS coordinates matches the Wasserstein
+distance between the distributions being embedded. Clusters are colored in the
+same manner as in Fig. \ref{fig:hclust} and labeled by the specific pattern that
 distinguishes the pasts in some of the clusters. Note that the clusters and
-labels are directly drawn from Fig. \ref{fig:hclust} for reference. They are
-not the result of the MDS algorithm itself. However, interactive plotting
-approaches may allow for similar exploration from these decompositions without
-the need for prior clustering.
+labels are directly drawn from Fig. \ref{fig:hclust} for reference. They are not
+the result of the MDS algorithm itself. However, interactive plotting approaches
+may allow for similar exploration from these decompositions without the need for
+prior clustering.
 
@@
 corresponds to a transient state induced by seeing all $1$s and then a $0$.
-The latter then synchronizes to right cluster.
+The latter then synchronizes to right cluster. By comparison, the fair coin's
+MDS embedding would have a single cluster since every past predicts the same
+future.
 
@@
 of the underlying dynamical system's (logistic map's) geometry.
+
+We also note that the direct reconstruction of the logistic map in this case is
+possible due to the choice of partition to transform the continuous variable $y$
+into the discrete variable $x$. The partition we chose is a \emph{generating
+partition} of the dynamical system \cite{Kurk03a}, which is a special kind of
+partition which preserves many of the measure-theoretic and topological features
+of the original system. Non-generating partitions would not yield the original
+logistic map if our methods are applied to them; however, nor would they if any
+other method were applied, as a non-generating partition represents an
+irreversible loss of information. In this case predictive state reconstruction
+would still recover the minimal sufficient statistics of the past sequence for
+future prediction (see, for instance, the discussion in \cite{Crut98d}).
 

### 2022-09-28 paper/manuscript.tex
JPC: Resubmission, cover letter, authors' response and manuscript editted.

@@
 
-\preprint{\arxiv{2206.XXXXX}}
+\preprint{\arxiv{2206.04198}}
 
@@
 then we take our data to be all subsequences of length $n+k$,
-splitting each subsequence into words of length $k$ and $n$ respectively. From
+splitting each subsequence into words of length $k$ and $n$, respectively. From
 these ``past/future'' pairs we can construct an empirical conditional
@@
 Examples 3 through 5, in particular, involve long-range
-correlations in their sequences which arise from structures which, upon
+correlations in their sequences that arise from structures which, upon
 ``opening,'' must be ``closed'' an arbitrary time later. Such structures are

### 2022-09-28 paper/publish.py

@@ -1,0 +1,555 @@
+#!/usr/bin/env python
+###############################################################################
+# publish.py
+#
+# A script to take a latex project and produce a version more suitable for
+# archaic journals.
+#
+# Copyright 2016 Ryan G. James.
+# This work may be used freely as long as this notice is included.
+# The work is provided "as is" without warranty, express or implied.
+###############################################################################
+
+try:
+    from halo import Halo
+    SPIN = True
+except:
+    SPIN = False
+import click
+import logbook
+from logbook.more import ColorizedStderrHandler
+try:
+    import pathlib2 as pathlib
+except:
+    import pathlib
+try:
+    import regex as re
+    re.DEFAULT_VERSION = re.VERSION1
+except:
+    import re
+import sarge
+import sys
+
+from itertools import product
+
+LOG_STREAM = ColorizedStderrHandler()
+LOG_STREAM.format_string = "\r{record.level_name}: {record.message}"
+LOG = logbook.Logger('Publisher')
+
+RUN_CMD = sarge.capture_both
+
+DIRECTORY_NAME = 'publish'
+
+TEMPLATE_FILES = [
+    'cmechabbrev.tex',
+    'cmvec.sty',
+    'dynlearn.sty',
+    'idiagrams.tikz',
+    'infotheory.tex',
+    'vaucanson.tikz',
+]
+
+UNWANTED_EXTENSIONS = [
+    'aux',
+    'bbl',
+    'bib',
+    'blg',
+    'brf',
+    'log',
+    'lox',
+    'out',
+    'sed',
+    'tikz',
+    'toc',
+]
+
+IMAGE_EXTENSIONS = [
+    'pdf',
+    'eps',
+    'ps',
+    'png',
+    'jpg',
+    'bmp',
+]
+
+
+@click.command()
+@click.argument('tex')
+@click.option('--inline', is_flag=True, help='inline all tex and sty files')
+@click.option('--debug', is_flag=True, help='show debug info')
+def publish(tex, inline, debug):
+    """
+    This command performs several operations:
+
+    - Produce a `publish` directory with tex files copied over
+
+    - Remove comments
+
+    - Factor out tikz code
+
+    - Rename images to figure-n.pdf and place them alongside the tex
+
+    - Factors out cref
+
[…]
